---
title: "modelInputData"
author: "Leticia Duarte"
output:
  html_notebook:
    toc: true
    toc_float: true
  html_document:
    toc: true
    df_print: paged
editor_options:
  chunk_output_type: inline
---

<style type="text/css">

body{ /* Normal  */
      font-size: 13px;
  }
td {  /* Table  */
  font-size: 12px;
}
h1.title {
  font-size: 20px;
  font-weight: bold;
}
h1 { /* Header 1 */
  font-size: 20px;
  font-weight: bold;
}
h2 { /* Header 2 */
    font-size: 17px;
  font-weight: bold;
}
h3 { /* Header 3 */
    font-size: 15px;
  font-weight: bold;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre{ /* Code block - determines code spacing between lines */
    font-size: 12px;
}
h1, .h1, h2, .h2, h3, .h3, h4, .h4 {
    margin-top: 24xpx;
}
</style>

# 1. Packages

```{r}
library(readr)
library(readxl)
library(dplyr)
library(tidyr)
library(stringr)
library(magrittr)
library(sf)
library(ggplot2)
library(cowplot)
library(remotes)
#remotes::install_github("Tong-Chen/YSX")
library("YSX")
library(patchwork)
```

# 2. Raw data

```{r}
# import data
data <-
  readr::read_csv(
    "rawData_new/FateWolvesTerritory1998_2022_250509.csv")

unique(data$fate) #alive, disapp, other, censored, natural

```

# 3. Adding variables

## 3.1. Inbreeding

```{r}
# calculate average and maximum inbreeding per territory
data_withFi <- 
  data %>%
  group_by(territory) %>%
  dplyr::mutate(inbreeding = as.numeric(inbreeding)) %>%
  dplyr::mutate(mean_inbreeding = mean(inbreeding),
         max_inbreeding = max(inbreeding))

# calculate new inbreeding value which considers the inbreeding value of the individual that disappeared
# in the case both individuals disappeared, the maximum inbreeding is used

data_withFi_new <- data_withFi %>%
  group_by(territory, autumn) %>%
  mutate(
    # extract inbreeding values of disappeared individuals in this group
    disappeared_inbr = list(inbreeding[fate == "disapp"]),
    # determine the desired new inbreeding value for this group based on conditions
    inbreeding_disapp = case_when(
      length(disappeared_inbr[[1]]) == 1 ~ disappeared_inbr[[1]][1],                  # exactly one disappeared
      length(disappeared_inbr[[1]]) > 1 ~ max(disappeared_inbr[[1]]),                 # both disappeared (max)
      TRUE ~ inbreeding                                                 # keep values for individuals that did not disappear
    )
  ) %>%
  ungroup() %>%
  select(-disappeared_inbr)  # remove intermediate column

```


## 3.2. Spatial data

```{r}
norway <-
  sf::st_read("spatialData/gadm36_NOR_shp/gadm36_NOR_2.shp")
sweden <-
  sf::st_read("spatialData/gadm36_SWE_shp/gadm36_SWE_2.shp")
scandinavia <- rbind(norway, sweden) %>%
  select(NAME_1, NAME_2, geometry)

# add municipality (gadm classification)
# convert to spatial data and add gadm names
dataToModel1 <-
  st_as_sf(
    x = data_withFi_new,
    #convert to spatial features (sf)
    coords = c("X_coordinate",
               "Y_coordinate"),
    crs = "EPSG:3021"
  ) %>% #RT90
  st_transform(crs = st_crs(scandinavia)) %>%
  st_intersection(scandinavia) %>% # add gadm names for municipalities
  st_drop_geometry() # back to data frame

```

## 3.3. Legally killed wolves

```{r}
# number of legally killed wolves
killedWolvesData <- readr::read_csv("rawData_new/Legal_wolf_harvest.csv") %>%
  separate(Dødsdato, c("DødsYear", "DødsMonth","DødsDay")) %>%
  dplyr::mutate(DødsYear = as.numeric(DødsYear),
         DødsMonth = as.numeric(DødsMonth),
         DødsDay = as.numeric(DødsDay)) %>%
  select("RovbaseID", "DødsYear",
         "Nord (RT90)", "Øst (RT90)") %>% # columns of interest 
  dplyr::filter(DødsYear > 1998) # remove records outside of the monitoring period

# convert to spatial data
killedWolvesSpatial <- st_as_sf(x = killedWolvesData,
                                coords = c("Øst (RT90)", "Nord (RT90)"),
                                crs = "EPSG:3021") %>% #RT90
  st_transform(crs = st_crs(scandinavia)) #%>% 
  #select(RovbaseID,geometry)

# intersect legally killed wolves with scandinavia admn areas
wolfLKill <- st_intersection(killedWolvesSpatial, scandinavia)

# count legal kills per municipality per year
wolfLKill_municipality <- wolfLKill %>% dplyr::count(DødsYear, NAME_2) %>%
  dplyr::rename(wolfLKill_n = n)


# build base dataframe 
kommun <- rep(levels(as.factor(dataToModel1$NAME_2)), # uses the kommunes with wolf territories
              each=length(c(2000:2022)))
# data starts in 2000 - see wolfLKill_municipality
year <- rep(c(2000:2022), length(levels(as.factor(dataToModel1$NAME_2))))
df <- data.frame(kommun = kommun,year = year)


# cumulative sum in the last five years include the target year and the previous 4
# cumulative slide was calculated based on https://stackoverflow.com/questions/72587840/how-to-get-the-cummean-of-the-last-n-rows-before-the-last-row-using-dplyr
wolfLKillAllMetrics <- left_join(df,
                                 wolfLKill_municipality,
                  by=c('kommun'='NAME_2', 'year'='DødsYear')) %>%
  #ungroup() %>%
  group_by(kommun) %>%
  # replaces NAs by zeros. This considers that records of wolf kills were recorded since 2000 (first year with records)
  dplyr::mutate_at(c('wolfLKill_n'), ~tidyr::replace_na(.,0)) %>%
  dplyr::mutate(wolfLKill_last5y = slider::slide_dbl(wolfLKill_n, sum, .before = 4, .after = 0),
         .after=year) %>%
  dplyr::mutate(wolfLKill_cumall = cumsum(wolfLKill_n),.after=year)


# join to dkommun# join to data to model  
dataToModel2 <- left_join(dataToModel1,
                          wolfLKillAllMetrics,
                          by=c('NAME_2'='kommun', 'autumn'='year')) %>%
  # replaces zeros to NAs before 1999 because there was no wolf kill data before 2000
  dplyr::mutate(
    wolfLKill_n = ifelse(autumn <= 1999, 0, wolfLKill_n),
    wolfLKill_cumall = ifelse(autumn <= 1999, 0, wolfLKill_cumall),
    wolfLKill_last5y = ifelse(autumn <= 1999, 0, wolfLKill_last5y))

```

## 3.4. Wolf attacks on sheep

```{r}
# import files for Norway and Sweden
sheep_data_sw <- readr::read_csv("rawData_new/CAWI27032025121630581_Sweden_250327_sheep.csv") %>%
  select("Fynddatum","Nord (RT90)","Öst (RT90)","Kommune","Number of injured or killed or missing sheep") %>%
  rename("NoAffectedSheep" = "Number of injured or killed or missing sheep")

sheep_data_nw <- readr::read_csv("rawData_new/CAWI27032025121531507_Norway_250327_sheep.csv") %>%
  select("Funnetdato","Nord (RT90)","Øst (RT90)","Kommune","Number of injured or killed or missing sheep")%>%
  rename_with(~ colnames(sheep_data_sw))

# merge dataframes
sheep_data <- rbind(sheep_data_sw, sheep_data_nw)

# group rows that represent the same event according to date and location
# this is done to calculate the number of attacks
sheep_data$coord <- paste0(sheep_data$`Nord (RT90)`, ",", sheep_data$`Öst (RT90)`)
sheep_attacks <- sheep_data %>%
  group_by(Fynddatum,coord) %>% # group by unique date and coordinates
  summarize(
    `Nord (RT90)` = first(`Nord (RT90)`),
    `Öst (RT90)` = first(`Öst (RT90)`),
    Kommune = first(Kommune), # keep the first
    NoAffectedSheep = sum(NoAffectedSheep)  # sum the number of attacks
  ) %>%
  ungroup() %>% # ungroup to return to a normal data frame
  select(-coord)

# clean data
killedsheepclean <- sheep_attacks %>%
  na.omit() %>%
  separate(Fynddatum, c("year","month","day")) %>% 
  dplyr::mutate(year = as.numeric(year)) %>%
  dplyr::filter(year >= 1998)

# read as spatial points and change coordinate to RT90
killedsheepSpatial <- st_as_sf(x = killedsheepclean,
                               coords = c("Öst (RT90)", "Nord (RT90)"),
                               crs = "EPSG:3021") %>% #RT90
  st_transform(crs = st_crs(scandinavia)) %>% 
  select(year, "NoAffectedSheep", geometry) 

# intersect sheep attacks with scandinavia admn areas
sheepAttacks <- st_intersection(killedsheepSpatial, scandinavia)

# count number of sheep killed  and number of attacks per municipality
sheepAffected_municipality <- sheepAttacks %>%
  dplyr::group_by(year, NAME_2) %>%
  dplyr::summarise(NoAffectedSheep = sum(NoAffectedSheep),
            n = n(),.groups = "drop") %>%
  dplyr::rename(sheepAttacks_n = n)

# build base dataframe 
kommun <- rep(levels(as.factor(dataToModel2$NAME_2)), # uses the kommunes with wolf territories
              each=length(c(1998:2022)))
year <- rep(c(1998:2022), length(levels(as.factor(dataToModel2$NAME_2))))
df <- data.frame(kommun = kommun,year = year)

# cumulative metrics
# cumulative sum in the last five years include the target year and the previous 4
# cumulative slide was calculated based on https://stackoverflow.com/questions/72587840/how-to-get-the-cummean-of-the-last-n-rows-before-the-last-row-using-dplyr
sheepAllMetrics <- left_join(df,
                             sheepAffected_municipality,
                  by=c('kommun'='NAME_2', 'year'='year')) %>%
  dplyr::group_by(kommun) %>%
  # replaces NAs by zeros. This considers that records of sheep events were recorded since 1995 (first year with records)
  mutate_at(c('NoAffectedSheep','sheepAttacks_n'), ~replace_na(.,0)) %>%
  dplyr::mutate(NoAffectedSheep_last5y = slider::slide_dbl(NoAffectedSheep, sum, .before = 4, .after = 0),
         .after=year) %>%
  dplyr::mutate(NoAffectedSheep_cumall = cumsum(NoAffectedSheep),.after=year)
 
# join to data to model  
dataToModel3 <- left_join(dataToModel2,
                      sheepAllMetrics,
                      by=c('NAME_2'='kommun', 'autumn'='year'))

```

## 3.5. Wolf attacks on dogs

```{r}
# import data for Norway and Sweden
dogData_sw <- readr::read_csv("rawData_new/CAWI27032025121630581_Sweden_250327_dogs.csv") %>%
  select("Fynddatum","Nord (RT90)","Öst (RT90)","Kommune","Number of injured or killed dogs") %>%
  rename("NoAffectedDogs" = "Number of injured or killed dogs")

dogData_nw <- readr::read_csv("rawData_new/CAWI27032025121531507_Norway_250327_dogs.csv") %>%
  select("Funnetdato","Nord (RT90)","Øst (RT90)","Kommune","Number of injured or killed dogs")%>%
  rename_with(~ colnames(dogData_sw))

# merge dataframes
dog_data <- rbind(dogData_sw, dogData_nw)

# calculate attacks
dog_data$coord <- paste0(dog_data$`Nord (RT90)`, ",", dog_data$`Öst (RT90)`)
dog_data <- dog_data %>%
  group_by(Fynddatum,coord) %>% # group by unique date and coordinates
  summarize(
    `Nord (RT90)` = first(`Nord (RT90)`),
    `Öst (RT90)` = first(`Öst (RT90)`),
    Kommune = first(Kommune), # keep the first
    NoAffectedDogs = sum(NoAffectedDogs)  # sum the number of attacks
  ) %>%
  ungroup() %>% # ungroup to return to a normal data frame
  select(-coord)

# clean data
AttacksDogclean <- dog_data %>%
  na.omit() %>%
  separate(Fynddatum, c("year","month","day")) %>% 
  dplyr::mutate(year = as.numeric(year)) %>%
  dplyr::filter(year >= 1998)

# read as spatial points and change coordinate to RT90
AttacksDogSpatial <- st_as_sf(x = AttacksDogclean,
                           coords = c("Öst (RT90)", "Nord (RT90)"),
                           crs = "EPSG:3021") %>% #RT90
  st_transform(crs = st_crs(scandinavia)) %>% 
  select(NoAffectedDogs, year, geometry)

# intersect attacks on dogs with scandinavia admn areas
AttackDogs <- st_intersection(AttacksDogSpatial, scandinavia)

# count number of attacks on dogs per year and per municipality
AttackDogs_municipality <- AttackDogs %>%
  group_by(year, NAME_2) %>%
  dplyr::summarise(NoAffectedDogs = sum(NoAffectedDogs),
                   n = n(),.groups = "drop") %>%
  dplyr::rename(AttackDogs_n = n)


# build base dataframe 
kommun <- rep(levels(as.factor(dataToModel3$NAME_2)),
              each=length(c(1998:2022)))
year <- rep(c(1998:2022), length(levels(as.factor(dataToModel3$NAME_2))))
df <- data.frame(kommun = kommun,year = year)

# cumulative metrics
# cumulative sum in the last five years include the target year and the previous 4
# cumulative slide was calculated based on https://stackoverflow.com/questions/72587840/how-to-get-the-cummean-of-the-last-n-rows-before-the-last-row-using-dplyr
dogAllMetrics <- left_join(df,
                           AttackDogs_municipality,
                           by=c('kommun'='NAME_2', 'year'='year')) %>%
  dplyr::group_by(kommun) %>%
  # replaces NAs by zeros. This considers that records of sheep events were recorded since 1995
  # (first year with records)
  mutate_at(c('NoAffectedDogs','AttackDogs_n'), ~replace_na(.,0)) %>%
  dplyr::mutate(NoAffectedDogs_last5y = slider::slide_dbl(NoAffectedDogs, sum, .before = 4, .after = 0),
         .after=year) %>%
  dplyr::mutate(NoAffectedDogs_cumall = cumsum(NoAffectedDogs),.after=year)

# join to data to model  
dataToModel4 <- left_join(dataToModel3,
                          dogAllMetrics,
                          by=c('NAME_2'='kommun', 'autumn'='year'))
```


## 3.6. Individual income

```{r}
# notes for norway data download

# Rømskog municipality in Norway no longer exists under that name.
# On January 1, 2020, Rømskog merged with Aurskog-Høland municipality
# in Viken County. So now, Rømskog is part of Aurskog-Høland.
# norway average gross individual income was extracted from https://www.ssb.no/en/statbank/table/05854/
# codes in kommun original were removed manually in excel, column names were also edited manually 


# Load the CSV file
income_norway <- read_delim("rawData_new/05854_MainEntriesFromTheTaxAssessmentForResidents17yearsAndOlder_individualIncomeNorway_edited.csv",
                 delim = ";", col_types = cols(.default = "c"))

# Delete the column 'kommun subset'
income_norway_filtered <- income_norway %>% select(-`kommun original`, -`age`, -`metric`)

# Convert all columns, except column "kommun", to numeric
income_norway_filtered <- income_norway_filtered %>%
  mutate(across(-kommun, ~ suppressWarnings(as.numeric(.))))

# Calculate row means and then summarize to one row per group
income_norway_sum <- income_norway_filtered %>%
  group_by(kommun) %>%
  summarize(across(where(is.numeric), sum, na.rm = TRUE))

# Replace zeros with NAs
income_norway_sum <- income_norway_sum %>%
  mutate(across(where(is.numeric), ~ na_if(.x, 0))) %>%
  mutate(country = "Norway") %>%
  select(country, everything())

# notes for sweden data download
# excel was manually edited to remove first row and last rows
# kommun name only available at the first row for each kommun and was manually inputed for the remaining ages
# Before 2003 Knivsta was part of Uppsala Municipality and hence the NA values 
# swedish income data was extracted from Total earned income for residents in Sweden throughout the year by region, gender, age and income class. Years 1999 - 2023
# https://www.statistikdatabasen.scb.se/pxweb/sv/ssd/START__HE__HE0110__HE0110A/SamForvInk1/
# kommun names for all the age groups were manually edited as the original table only had them for the first line of each kommun 
# needed to work in excel at this step since the kommun names in Swedish were not reading correctly when trying with csv

# Load the Excel file
income_sweden <- read_excel("rawData_new/Sweden_HE0110J7_20250301-210646_edited.xlsx")

# Remove columns
income_sweden <- income_sweden %>% select(-age, -metric)

# Convert all columns, except column "kommun", to numeric
income_sweden <- income_sweden %>%
  mutate(across(-kommun, ~ suppressWarnings(as.numeric(.))))

# Calculate row means and then summarize to one row per group
income_sweden_avg <- income_sweden %>%
  group_by(kommun) %>%
  summarize(across(where(is.numeric), mean))

# Multiply all numeric values by a thousand and reduce decimal places to zero
# This is because the values in the Excel file are in K SEK
income_sweden_avg <- income_sweden_avg %>%
  mutate(across(where(is.numeric), ~ round(.x * 1000, 0))) %>%
  mutate(country = "Sweden") %>%
  select(country, everything())

# notes merging the two dataframes
# I have ignored that norway data is in NOK and swedish data is in SEK

# Merge df2_avg and df_sum by columns with the same name
income_merged_df <- full_join(income_norway_sum, income_sweden_avg, 
                       by = intersect(names(income_norway_sum), names(income_sweden_avg)))

# Calculate the average of each year
yearly_avg <- income_merged_df %>%
  summarize(across(where(is.numeric), mean, na.rm = TRUE))

# Divide each cell of a given year by the mean of the same year
normalized_df <- income_merged_df %>%
  mutate(across(where(is.numeric), ~ . / yearly_avg[[cur_column()]]))

# Save the merged dataframe to a new CSV file
# write_csv(normalized_df, "data/rawData/income_forWolf/Merged_Data.csv")

# Pivot normalized_df to have columns: country, kommun, year, and value
pivoted_df <- normalized_df %>%
  pivot_longer(cols = -c(country, kommun), names_to = "year", values_to = "income") %>%
  dplyr::mutate(year = as.numeric(year))

# join to data to model
dataToModel5 <-
 left_join(
   dataToModel4,
   pivoted_df,
by = c('NAME_2' = 'kommun', 'autumn' = 'year')
   )

# check NA values
# select rows with NA income values
na_rows <- dataToModel5 %>%
  select('NAME_2','income','autumn') %>%
  filter(is.na(income))

unique(na_rows$NAME_2)

# check manually problems with these kommune names
# some NA values were due to missing data in raw data (no data for 1998)
# some were due to different kommune names:
# AurskogHøland --> Aurskog-Høland
# SørOdal --> Sør-Odal
# StorElvdal --> Stor-Elvdal
# Malung-Sälen --> Malung
# VålerHedmark and VålerInndlandet --> Våler
# Rømskog is Aurskog-Høland since 2020 - but in income dataset it still has the same name

pivoted_df <- pivoted_df %>%
  na.omit(kommun) %>%
  mutate(
    kommun = str_replace_all(kommun, "AurskogHøland", "Aurskog-Høland"),
    kommun = str_replace_all(kommun, "SørOdal", "Sør-Odal"),
    kommun = str_replace_all(kommun, "StorElvdal", "Stor-Elvdal"),
    kommun = str_replace_all(kommun, "Malung-Sälen", "Malung"),
    kommun = str_replace_all(kommun, "VålerHedmark", "Våler"),
    kommun = str_replace_all(kommun, "VålerInnlandet", "Våler"))

dataToModel4_v2 <- dataToModel4 %>%
  mutate(
    NAME_2 = ifelse(NAME_2 == "Aurskog-Høland" & autumn >= 2020, "Rømskog", NAME_2)
  )


# join to data to model
dataToModel5 <-
 left_join(
   dataToModel4,
   pivoted_df,
by = c('NAME_2' = 'kommun', 'autumn' = 'year')
   )

```


## 3.7. Prepare the data

```{r}
# add years since first detection for test with survival analyses
dataToModel6 <- 
  dataToModel5 %>%
  group_by(territory) %>%
  dplyr::mutate(
    first_yeardet = min(autumn),
    last_yeardet = max(autumn),
    years_firstdet = autumn - first_yeardet,
    time_start = autumn - first_yeardet,
    time_end = spring - first_yeardet
  ) 

# add population size
# note: in this way legally removed territories in one year or territories that
# disappeared due to natural causes are still considered in the census of that
# spring. The same for other causes of death 
wolfPopSize <-
  dataToModel6 %>%
  group_by(autumn) %>%
  dplyr::summarise(
    Spring = unique(autumn),
    # number of unique territories per year, independently of their fate
    wolfPopSize = length(unique(territory))
  )

dataToModel7 <-
  left_join(
    dataToModel6,
    wolfPopSize,
    by = c('autumn' = 'autumn')
  )

#compiled_dataset <- dataToModel7 %>%
#  select(1:39,41:44,46:49,52,59)



#writexl::write_xlsx(compiled_dataset,"rawData_new/compiled_dataset.xlsx")

```


# 4. Model input data

## 4.1. Territory level

```{r}
unique(dataToModel7$fate) #"alive"    "disapp"   "censored" "other"

territoriesToModel <- dataToModel7 %>%
  filter(!any(fate %in% "other")) %>% #exclude fate "other"
  mutate(fate_binary = ifelse(fate == "disapp", 1, 0),.after=fate) %>% #create fate_binary variable, where "disapp" = 1, and "alive" and "censored" = 0
  arrange(territory, autumn)

length(unique(territoriesToModel$territory)) # 372
nrow(territoriesToModel) # 1558 only one individual per territory
unique(territoriesToModel$fate)

#make dataset clean
territoriesToModel2 <- territoriesToModel[,-c(42,47,52,53,60), drop=FALSE]

# count disappearances and ongoing territories per year ------------------------
df <- territoriesToModel2 %>% 
  group_by(autumn) %>%
  dplyr::summarise(
    ongoing = sum(fate %in% c("alive","censored")),
    disappearance = sum(fate %in% "disapp"),
    sum = ongoing + disappearance,
    prop_disappearance = round(disappearance/sum, 2),
    n()
  )

sum(df$disappearance) # 419 # double checking
sum(df$ongoing) # 1139 # double checking
print(df, n=30)

# the plot
ggplot(data=df, aes(x=autumn, y=prop_disappearance, group=1)) +
  geom_line()+
  geom_point()

# model data scaled ------------------------------------------------------------
modeldatascaled <- territoriesToModel2
modeldatascaled$ruggedness_sc <- scale(modeldatascaled$ruggedness)[,1]
modeldatascaled$gravel_sc <- scale(modeldatascaled$gravel)[,1]
modeldatascaled$paved_sc <- scale(modeldatascaled$paved)[,1]
modeldatascaled$human_density_sc <- scale(modeldatascaled$human_density)[,1]               
modeldatascaled$snow_cover_sc <- scale(modeldatascaled$snow_cover)[,1]
modeldatascaled$moose_harvest_density_sc <- scale(modeldatascaled$moose_harvest_density)[,1]  
modeldatascaled$number_neighbour_terr_sc <- scale(modeldatascaled$number_neighbour_terr)[,1]
modeldatascaled$bear_density_shot_sc <- scale(modeldatascaled$bear_density_shot)[,1]
modeldatascaled$bear_density_live_sc <- scale(modeldatascaled$bear_density_live)[,1]
modeldatascaled$night_light_sc <- scale(modeldatascaled$night_light)[,1]
modeldatascaled$elevation_mean_sc <- scale(modeldatascaled$elevation_mean)[,1]
modeldatascaled$elevation_stdv_sc <- scale(modeldatascaled$elevation_stdv)[,1]
modeldatascaled$inbreeding_sc <- scale(modeldatascaled$inbreeding)[,1] 
modeldatascaled$mean_inbreeding_sc <- scale(modeldatascaled$mean_inbreeding)[,1] 
modeldatascaled$max_inbreeding_sc <- scale(modeldatascaled$max_inbreeding)[,1]
modeldatascaled$inbreeding_disapp_sc <- scale(modeldatascaled$inbreeding_disapp)[,1]
modeldatascaled$wolfLKill_cumall_sc <- scale(modeldatascaled$wolfLKill_cumall)[,1]  
modeldatascaled$wolfLKill_last5y_sc <- scale(modeldatascaled$wolfLKill_last5y)[,1]     
modeldatascaled$wolfLKill_n_sc <- scale(modeldatascaled$wolfLKill_n)[,1]  
modeldatascaled$NoAffectedSheep_cumall_sc <- scale(modeldatascaled$NoAffectedSheep_cumall)[,1]
modeldatascaled$NoAffectedSheep_last5y_sc <- scale(modeldatascaled$NoAffectedSheep_last5y)[,1]
modeldatascaled$NoAffectedSheep_sc <- scale(modeldatascaled$NoAffectedSheep)[,1]        
modeldatascaled$sheepAttacks_n_sc <- scale(modeldatascaled$sheepAttacks_n)[,1]  
modeldatascaled$NoAffectedDogs_cumall_sc <- scale(modeldatascaled$NoAffectedDogs_cumall)[,1]  
modeldatascaled$NoAffectedDogs_last5y_sc <- scale(modeldatascaled$NoAffectedDogs_last5y)[,1]
modeldatascaled$NoAffectedDogs_sc <- scale(modeldatascaled$NoAffectedDogs)[,1]
modeldatascaled$AttackDogs_n_sc <- scale(modeldatascaled$AttackDogs_n)[,1]          
modeldatascaled$income_sc <- scale(modeldatascaled$income)[,1] 
modeldatascaled$pop_size_territories_sc <- scale(modeldatascaled$pop_size_territories)[,1]

df.territory <- as_tibble(modeldatascaled)

# territory data
write.csv(df.territory,'modelInputData_new/df.territory_new.csv')
writexl::write_xlsx(df.territory,"modelInputData_new/df.territory_new.xlsx")

```


### 4.1.1. Correlations

```{r}
## correlations with all variables ---------------------------------------------
colnames(territoriesToModel2)
covToCorrelate_terr <- territoriesToModel2[c("wolfPopSize","ruggedness","paved","gravel","snow_cover","human_density","moose_harvest_density","bear_density_shot","bear_density_live","night_light","elevation_mean","elevation_stdv","inbreeding","number_neighbour_terr","mean_inbreeding","max_inbreeding","inbreeding_disapp","wolfLKill_cumall","wolfLKill_last5y","wolfLKill_n","NoAffectedSheep_cumall","NoAffectedSheep_last5y","NoAffectedSheep","sheepAttacks_n","NoAffectedDogs_cumall","NoAffectedDogs_last5y","NoAffectedDogs","AttackDogs_n","income")]

cormat <- round(
  cor(
    covToCorrelate_terr,
    use = "pairwise.complete.obs",
    method = "spearman"),
  2)

cormat
upper_tri <- get_upper_tri(cormat)
melted_cormat <- reshape2::melt(upper_tri, na.rm = TRUE)

p1_terr <- ggplot(melted_cormat, aes(Var2, Var1, fill = value))+
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  ggtitle(label = "Spearman correlation - territory level - all covariates") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1,
                                   hjust = 1, size = 8)) +
  theme(axis.text.y = element_text(size = 8)) +
  geom_text(aes(Var2, Var1, label = value), color = "white", size = 3)

print(p1_terr)

# highlyCorrelatedVars <- melted_cormat %>%
#   filter(value >0.5 | value < -0.5)
#highlyCorrelatedVars


# subset the covariates to keep
subset_keep <- territoriesToModel[c("wolfPopSize","ruggedness","paved","gravel","snow_cover","human_density","moose_harvest_density","bear_density_live","night_light","elevation_stdv","number_neighbour_terr","inbreeding_disapp","wolfLKill_last5y","NoAffectedSheep_last5y","NoAffectedDogs_last5y","income")]

# change covariates names
colnames(subset_keep) <- c("WolfPopSize","TerrRug","PavedRoad","GravelRoad",
                           "SnowCover","HumanDens","MooseHarv","BearDens",
                           "NightLight","ElevationSTDV","WolfNeighbourTerr",
                           "Inbreeding","WolfHarv_last5y","SheepAttacks_last5y",
                           "DogAttacks_last5y","Income")


## correlation between covariates representing the same mechanism --------------
VarsToTest <- c("TerrRug",
                "PavedRoad",
                "GravelRoad",
                "SnowCover",
                "HumanDens",
                "NightLight",
                "ElevationSTD")

VarsToTest <- c("DogAttacks_last5y",
                "SheepAttacks_last5y",
                "WolfHarv_last5y",
                "MooseHarv",
                "Income")  

VarsToTest <- c("BearDens",
                "WolfNeighbourTerr")

sub_VarsToTest <- subset_keep %>%
  select(all_of(VarsToTest))

sub_cormat <- round(
  cor(
    sub_VarsToTest,
    use = "pairwise.complete.obs",
    method = "spearman"),2
)

sub_upper_tri <- get_upper_tri(sub_cormat)
sub_melted_cormat <-
  reshape2::melt(sub_upper_tri, na.rm = TRUE)

p2_sub <- ggplot(sub_melted_cormat, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  ggtitle(label = "Spearman correlation - Territory level - Competition") +
  #change title depending on the hypothesis
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1,
    size = 16
  )) +
  theme(axis.text.y = element_text(size = 16)) +
  geom_text(aes(Var2, Var1, label = value),
            color = "white",
            size = 6)
p2_sub


## correlations between chosen variables ---------------------------------------

sub_cormat3 <- round(
  cor(
    subset_keep,
    use = "pairwise.complete.obs",
    method = "spearman"),2
)

sub_upper_tri3 <- get_upper_tri(sub_cormat3)
sub_melted_cormat3 <-
  reshape2::melt(sub_upper_tri3, na.rm = TRUE)

p3_sub <- ggplot(sub_melted_cormat3, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  ggtitle(label = "Spearman correlation - Territory level") +
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1,
    size = 16
  )) +
  theme(axis.text.y = element_text(size = 14)) +
  geom_text(aes(Var2, Var1, label = value),
            color = "white",
            size = 5)
p3_sub



## correlations of final covariates --------------------------------------------

subset_keep2 <- subset_keep[c("WolfPopSize","TerrRug","GravelRoad",
                              "ArtificialArea","HumanDens",
                              "DogAttacks_last5y","SheepAttacks_last5y",
                              "LegWolfHarv_last5y","IncomeProp",
                              "BearHarv","WolfNeighbourTerr","Mean_InbreedCoef")]

sub_cormat4 <- round(
  cor(
    subset_keep2,
    use = "pairwise.complete.obs",
    method = "spearman"),2
)

sub_upper_tri4 <- get_upper_tri(sub_cormat4)
sub_melted_cormat4 <-
  reshape2::melt(sub_upper_tri4, na.rm = TRUE)

p5_sub <- ggplot(sub_melted_cormat4, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  ggtitle(label = "Spearman correlation - Territory level - Final variables") +
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1,
    size = 16
  )) +
  theme(axis.text.y = element_text(size = 16)) +
  geom_text(aes(Var2, Var1, label = value),
            color = "white",
            size = 6)
p5_sub

```

### 4.1.2. Calculate min, max and mean

```{r}
min_cov <- format(sapply(df.territory,min, na.rm = TRUE), scientific = FALSE)
min_cov <- as.data.frame(min_cov)

max_cov <- format(sapply(df.territory,max, na.rm = TRUE), scientific = FALSE)
max_cov <- as.data.frame(max_cov)

mean_cov <- format(sapply(df.territory,mean, na.rm = TRUE), scientific = FALSE)
mean_cov <- as.data.frame(mean_cov)

var <- data.frame(min_cov, max_cov, mean_cov)

var <- var[c("wolfPopSize","ruggedness","paved","gravel","human_density","snow_cover","night_light","NoAffectedSheep_last5y","NoAffectedDogs_last5y","moose_harvest_density","income","wolfLKill_last5y","inbreeding","number_neighbour_terr","bear_density_live"),]

write.csv(var, "ModelInputData_new/info_variables_territory.csv")

```

## 4.2. Population level

```{r}
# summarise territory data per year
# adopted from territory selection but includes all fates to calculate pop.size
# slice_max is used to select only one row per territory per year
# all variables are the same for both individuals except for inbreeding
# however, with inbreeding_disapp we also have the same inbreeding value for both individuals; thus, the individual chosen is irrelevant and random

yearlyTerritoryStats <- 
  dataToModel7 %>%
  group_by(territory, autumn) %>%
  slice_max(inbreeding_disapp, with_ties = FALSE) %>% #returns the first row when value is the same
  ungroup() %>%
  arrange(territory, autumn) %>%
  dplyr::mutate(year = autumn) %>%
  group_by(year) %>%
  dplyr::summarise(
    wolfPopSize = length(unique(territory)),
    disappearance = sum(fate == "disapp", na.rm = TRUE),
    n_territories_without_disap = wolfPopSize - disappearance,
    prop_disappearance = disappearance / wolfPopSize,
  )
#head(yearlyTerritoryStats, n=30)


yearlyTerritoryFeatures <- territoriesToModel2 %>%
  dplyr::mutate(year = autumn) %>%
  group_by(year) %>%
  #dplyr::filter(Fate_binary == 1) %>% # filter only disappearances
  dplyr::summarise(
    ruggedness = mean(ruggedness, na.rm = TRUE),
    gravel = mean(gravel, na.rm = TRUE),
    paved = mean(paved, na.rm = TRUE),
    human_density = mean(human_density, na.rm = TRUE),
    snow_cover = mean(snow_cover, na.rm = TRUE),
    night_light = mean(night_light, na.rm = TRUE),
    elevation_mean = mean(elevation_mean, na.rm = TRUE),
    elevation_stdv = mean(elevation_stdv, na.rm = TRUE),
    moose_harvest_density = mean(moose_harvest_density, na.rm = TRUE),
    mean_inbreeding = mean(mean_inbreeding, na.rm = TRUE),
    max_inbreeding = mean(max_inbreeding, na.rm = TRUE),
    inbreeding_disapp = mean(inbreeding_disapp, na.rm = TRUE),
    number_neighbour_terr = mean(number_neighbour_terr, na.rm = TRUE),
    bear_density_shot = mean(bear_density_shot, na.rm = TRUE),
    bear_density_live = mean(bear_density_live, na.rm = TRUE),
    wolfLKill_n = sum(wolfLKill_n, na.rm = TRUE),
    wolfLKill_cumall = sum(wolfLKill_last5y, na.rm = TRUE),
    wolfLKill_last5y = sum(wolfLKill_last5y, na.rm = TRUE),
    sheepAttacks_n = sum(sheepAttacks_n, na.rm = TRUE),
    NoAffectedSheep_last5y = sum(NoAffectedSheep_last5y, na.rm = TRUE),
    NoAffectedSheep_cumall = sum(NoAffectedSheep_cumall, na.rm = TRUE),
    NoAffectedSheep = sum(NoAffectedSheep, na.rm = TRUE),
    AttackDogs_n = sum(AttackDogs_n, na.rm = TRUE),
    NoAffectedDogs_last5y = sum(NoAffectedDogs_last5y, na.rm = TRUE),
    NoAffectedDogs_cumall = sum(NoAffectedDogs_cumall, na.rm = TRUE),
    NoAffectedDogs = mean(NoAffectedDogs, na.rm = TRUE),
    income = mean(income, na.rm = TRUE),
    n = n()
  ) %>%
  arrange(year)

df.year <- left_join(yearlyTerritoryStats,
                     yearlyTerritoryFeatures,
                     by = c('year' = 'year'))

# scale covariates
df.year$ruggedness_sc <- scale(df.year$ruggedness)[,1]
df.year$gravel_sc <- scale(df.year$gravel)[,1]
df.year$paved_sc <- scale(df.year$paved)[,1]
df.year$human_density_sc <- scale(df.year$human_density)[,1]               
df.year$snow_cover_sc <- scale(df.year$snow_cover)[,1]
df.year$moose_harvest_density_sc <- scale(df.year$moose_harvest_density)[,1]  
df.year$number_neighbour_terr_sc <- scale(df.year$number_neighbour_terr)[,1]
df.year$bear_density_shot_sc <- scale(df.year$bear_density_shot)[,1]
df.year$bear_density_live_sc <- scale(df.year$bear_density_live)[,1]
df.year$night_light_sc <- scale(df.year$night_light)[,1]
df.year$elevation_mean_sc <- scale(df.year$elevation_mean)[,1]
df.year$elevation_stdv_sc <- scale(df.year$elevation_stdv)[,1]
df.year$inbreeding_sc <- scale(df.year$inbreeding_disapp)[,1] 
df.year$mean_inbreeding_sc <- scale(df.year$mean_inbreeding)[,1] 
df.year$max_inbreeding_sc <- scale(df.year$max_inbreeding)[,1]   
df.year$inbreeding_disapp_sc <- scale(df.year$inbreeding_disapp)[,1]
df.year$wolfLKill_cumall_sc <- scale(df.year$wolfLKill_cumall)[,1]  
df.year$wolfLKill_last5y_sc <- scale(df.year$wolfLKill_last5y)[,1]     
df.year$wolfLKill_n_sc <- scale(df.year$wolfLKill_n)[,1]  
df.year$NoAffectedSheep_cumall_sc <- scale(df.year$NoAffectedSheep_cumall)[,1]
df.year$NoAffectedSheep_last5y_sc <- scale(df.year$NoAffectedSheep_last5y)[,1]
df.year$NoAffectedSheep_sc <- scale(df.year$NoAffectedSheep)[,1]        
df.year$sheepAttacks_n_sc <- scale(df.year$sheepAttacks_n)[,1]  
df.year$NoAffectedDogs_cumall_sc <- scale(df.year$NoAffectedDogs_cumall)[,1]  
df.year$NoAffectedDogs_last5y_sc <- scale(df.year$NoAffectedDogs_last5y)[,1]
df.year$NoAffectedDogs_sc <- scale(df.year$NoAffectedDogs)[,1]
df.year$AttackDogs_n_sc <- scale(df.year$AttackDogs_n)[,1]          
df.year$income_sc <- scale(df.year$income)[,1] 
df.year$wolfPopSize_sc <- scale(df.year$wolfPopSize)[,1]

write_csv2(df.year,
          'modelInputData_new/df.year.csv')

writexl::write_xlsx(df.year,"modelInputData_new/df.year_new.xlsx")

```

### 4.2.1. Correlations

```{r}
# Note: covariate correlation had to be calculated for the different scales as big discrepancies between scales were detected
colnames(df.year)
covToCorrelate_pop <- df.year[c("wolfPopSize","ruggedness","paved","gravel","snow_cover","human_density","moose_harvest_density","bear_density_shot","bear_density_live","night_light","elevation_mean","elevation_stdv","number_neighbour_terr","mean_inbreeding","max_inbreeding","inbreeding_disapp","wolfLKill_cumall","wolfLKill_last5y","wolfLKill_n","NoAffectedSheep_cumall","NoAffectedSheep_last5y","NoAffectedSheep","sheepAttacks_n","NoAffectedDogs_cumall","NoAffectedDogs_last5y","NoAffectedDogs","AttackDogs_n","income")]

cormat <- round(
  cor(
    covToCorrelate_pop,
    use = "pairwise.complete.obs",
    method = "spearman"),
  2)

cormat
upper_tri <- get_upper_tri(cormat)
melted_cormat <- reshape2::melt(upper_tri, na.rm = TRUE)

p1_pop <- ggplot(melted_cormat, aes(Var2, Var1, fill = value))+
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  ggtitle(label = "Spearman correlation - population level - all covariates") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1,
                                   hjust = 1, size = 8)) +
  theme(axis.text.y = element_text(size = 8)) +
  geom_text(aes(Var2, Var1, label = value), color = "white", size = 3)
p1_pop  


# subset the covariates to keep
subset_keep <- df.year[c("wolfPopSize","ruggedness","paved","gravel","snow_cover","human_density","moose_harvest_density","bear_density_live","night_light","elevation_stdv","number_neighbour_terr","inbreeding_disapp","wolfLKill_last5y","NoAffectedSheep_last5y","NoAffectedDogs_last5y","income")]

# change covariates names
colnames(subset_keep) <- c("WolfPopSize","TerrRug","PavedRoad","GravelRoad",
                           "SnowCover","HumanDens","MooseHarv","BearDens",
                           "NightLight","ElevationSTDV","WolfNeighbourTerr",
                           "Inbreeding","WolfHarv_last5y","SheepAttacks_last5y",
                           "DogAttacks_last5y","Income")


## correlation between covariates representing the same mechanism --------------

VarsToTest2 <- c("TerrRug",
                "PavedRoad",
                "GravelRoad",
                "SnowCover",
                "HumanDens",
                "NightLight",
                "ElevationSTDV")

VarsToTest2 <- c("DogAttacks_last5y",
                "SheepAttacks_last5y",
                "WolfHarv_last5y",
                "MooseHarv",
                "Income")  

VarsToTest2 <- c("BearDens",
                 "WolfNeighbourTerr")

sub_VarsToTest2 <- subset_keep %>%
  select(all_of(VarsToTest2))

sub_cormat2 <- round(
  cor(
    sub_VarsToTest2,
    use = "pairwise.complete.obs",
    method = "spearman"),2
)

sub_upper_tri2 <- get_upper_tri(sub_cormat2)
sub_melted_cormat2 <-
  reshape2::melt(sub_upper_tri2, na.rm = TRUE)

p3_sub <- ggplot(sub_melted_cormat2, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  ggtitle(label = "Spearman correlation - Population level - Socio-economic") +
  #change title depending on the hypothesis
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1,
    size = 16
  )) +
  theme(axis.text.y = element_text(size = 16)) +
  geom_text(aes(Var2, Var1, label = value),
            color = "white",
            size = 6)
p3_sub

## correlations between chosen variables ---------------------------------------

sub_cormat3 <- round(
  cor(
    subset_keep,
    use = "pairwise.complete.obs",
    method = "spearman"),2
)

sub_upper_tri3 <- get_upper_tri(sub_cormat3)
sub_melted_cormat3 <-
  reshape2::melt(sub_upper_tri3, na.rm = TRUE)

p4_sub <- ggplot(sub_melted_cormat3, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  ggtitle(label = "Spearman correlation - Population level") +
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1,
    size = 16
  )) +
  theme(axis.text.y = element_text(size = 14)) +
  geom_text(aes(Var2, Var1, label = value),
            color = "white",
            size = 5)
p4_sub


## correlations of final covariates --------------------------------------------

subset_keep2 <- subset_keep[c("WolfPopSize","GravelRoad","SnowCover","TerrRug",
                              "HumanDens","Mean_InbreedCoef")]

sub_cormat4 <- round(
  cor(
    subset_keep2,
    use = "pairwise.complete.obs",
    method = "spearman"),2
)

sub_upper_tri4 <- get_upper_tri(sub_cormat4)
sub_melted_cormat4 <-
  reshape2::melt(sub_upper_tri4, na.rm = TRUE)

p5_sub <- ggplot(sub_melted_cormat4, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  ggtitle(label = "Spearman correlation - Population level - Final variables") +
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1,
    size = 16
  )) +
  theme(axis.text.y = element_text(size = 16)) +
  geom_text(aes(Var2, Var1, label = value),
            color = "white",
            size = 6)
p5_sub

```

### 4.2.2. Calculate min, max and mean

```{r}

min_cov <- format(sapply(df.year,min, na.rm = TRUE), scientific = FALSE)
min_cov <- as.data.frame(min_cov)

max_cov <- format(sapply(df.year,max, na.rm = TRUE), scientific = FALSE)
max_cov <- as.data.frame(max_cov)

mean_cov <- format(sapply(df.year,mean, na.rm = TRUE), scientific = FALSE)
mean_cov <- as.data.frame(mean_cov)

var <- data.frame(min_cov, max_cov, mean_cov)

var <- var[c("wolfPopSize","ruggedness","paved","gravel","human_density","snow_cover","night_light","NoAffectedSheep_last5y","NoAffectedDogs_last5y","moose_harvest_density","income","wolfLKill_last5y","inbreeding_disapp","number_neighbour_terr","bear_density_live"),]

write.csv(var, "modelInputData_new/info_variables_population.csv")

```

### 4.2.3. Plots of variables correlated with WolfPopSize

```{r}
# calculate coefficients for each variable

# format
max_cov_t <- as.data.frame(t(max_cov))
str(max_cov_t)
max_cov_t[c(1:62)] <- as.numeric(max_cov_t[c(1:62)])

# coefficients
coeff_attacksDogs <- max_cov_t$wolfPopSize/max_cov_t$NoAffectedDogs_last5y

coeff_attacksSheep <- max_cov_t$wolfPopSize/max_cov_t$NoAffectedSheep_last5y

coeff_wolvesKilled <- max_cov_t$wolfPopSize/max_cov_t$wolfLKill_last5y

coeff_moose <- max_cov_t$wolfPopSize/max_cov_t$moose_harvest_density

coeff_income <- max_cov_t$wolfPopSize/max_cov_t$income

coeff_bear <- max_cov_t$wolfPopSize/max_cov_t$bear_density_live

coeff_wolfNeigh <- max_cov_t$wolfPopSize/max_cov_t$number_neighbour_terr

coeff_terrRug <- max_cov_t$wolfPopSize/max_cov_t$ruggedness

# add corrected wolfpopsize columns

df.year_corrected <- left_join(df.year, forfig2, by = "year")

```


```{r}
# plots
# DISCRETE
plot_wolvesKilled <- ggplot(df.year_corrected) + 
  
  # Manually set linetypes for the two categories and define their labels
  scale_linetype_manual(
    name = "", 
    values = c("Wolf population size (corrected)" = 1, 
               "Wolf population size (from reports)" = 2),
    labels = c("Wolf population size (corrected)", 
               "Wolf population size (from reports)")
  ) +
  
  geom_col(aes(x = year, y = wolfLKill_last5y, group = 1, fill = "#660000")) +
  
  geom_line(
    aes(
      x = year, 
      y = wolfPopSize_correcteddata / coeff_wolvesKilled, 
      linetype = "Wolf population size (corrected)"
    ), 
    colour = "black"
  ) +
  
  geom_line(
    aes(
      x = year, 
      y = wolfPopSize_NOTcorrecteddata_from_annual_reports / coeff_wolvesKilled, 
      linetype = "Wolf population size (from reports)"
    ), 
    colour = "darkgrey"
  ) +
  
  scale_y_continuous(
    name = "Number of wolves killed",
    sec.axis = sec_axis(
      ~ .*coeff_wolvesKilled, 
      name = "Wolf population size", 
      breaks = seq(0, 80, by = 10)
    )
  ) +
  
  scale_fill_manual('', values = '#660000', label = 'Number of wolves legally killed (5 years)') +
  
  theme_light() +
  
  theme(
    axis.title.y = element_text(size = 18),
    axis.title.y.right = element_text(size = 18),
    axis.title.x = element_text(size = 18),
    legend.position = "bottom",
    legend.box.margin = margin(-12, 0, -5, 0),
    legend.text = element_text(size = 18),
    axis.text = element_text(size = 16)
  ) +
  
  xlab("Year")

###
plot_dogs <- ggplot(df.year_corrected) + 
  
  # Manually set linetypes for the two categories and define their labels
  scale_linetype_manual(
    name = "", 
    values = c("Wolf population size (corrected)" = 1, 
               "Wolf population size (from reports)" = 2),
    labels = c("Wolf population size (corrected)", 
               "Wolf population size (from reports)")
  ) +
  
  geom_col(aes(x = year, y = NoAffectedDogs_last5y, group = 1, fill = "tan2")) +
  
  geom_line(
    aes(
      x = year, 
      y = wolfPopSize_correcteddata / coeff_attacksDogs, 
      linetype = "Wolf population size (corrected)"
    ), 
    colour = "black"
  ) +
  
  geom_line(
    aes(
      x = year, 
      y = wolfPopSize_NOTcorrecteddata_from_annual_reports / coeff_attacksDogs, 
      linetype = "Wolf population size (from reports)"
    ), 
    colour = "darkgrey"
  ) +
  
  scale_y_continuous(
    name = "Number of dogs attacked",
    sec.axis = sec_axis(
      ~ .*coeff_attacksDogs, 
      name = "Wolf population size", 
      breaks = seq(0, 80, by = 10)
    )
  ) +
  
  scale_fill_manual('', values = 'tan2', label = 'Number of dogs attacked (5 years)') +
  
  theme_light() +
  
  theme(
    axis.title.y = element_text(size = 18),
    axis.title.y.right = element_text(size = 18),
    axis.title.x = element_text(size = 18),
    legend.position = "bottom",
    legend.box.margin = margin(-12, 0, -5, 0),
    legend.text = element_text(size = 18),
    axis.text = element_text(size = 16)
  ) +
  
  xlab("Year")

###
plot_sheep <- ggplot(df.year_corrected) + 
  
  # Manually set linetypes for the two categories and define their labels
  scale_linetype_manual(
    name = "", 
    values = c("Wolf population size (corrected)" = 1, 
               "Wolf population size (from reports)" = 2),
    labels = c("Wolf population size (corrected)", 
               "Wolf population size (from reports)")
  ) +
  
  geom_col(aes(x = year, y = NoAffectedSheep_last5y, group = 1, fill = "sienna3")) +
  
  geom_line(
    aes(
      x = year, 
      y = wolfPopSize_correcteddata / coeff_attacksSheep, 
      linetype = "Wolf population size (corrected)"
    ), 
    colour = "black"
  ) +
  
  geom_line(
    aes(
      x = year, 
      y = wolfPopSize_NOTcorrecteddata_from_annual_reports / coeff_attacksSheep, 
      linetype = "Wolf population size (from reports)"
    ), 
    colour = "darkgrey"
  ) +
  
  scale_y_continuous(
    name = "Number of sheep attacked",
    sec.axis = sec_axis(
      ~ .*coeff_attacksSheep, 
      name = "Wolf population size", 
      breaks = seq(0, 80, by = 10)
    )
  ) +
  
  scale_fill_manual('', values = 'sienna3', label = 'Number of sheep attacked (5 years)') +
  
  theme_light() +
  
  theme(
    axis.title.y = element_text(size = 18),
    axis.title.y.right = element_text(size = 18),
    axis.title.x = element_text(size = 18),
    legend.position = "bottom",
    legend.box.margin = margin(-12, 0, -5, 0),
    legend.text = element_text(size = 18),
    axis.text = element_text(size = 16)
  )+
  
  xlab("Year")

###
# CONTINUOUS

plot_income <- ggplot(df.year_corrected, aes(x = year)) +
  
  geom_line(aes(y = income, colour = "Average annual income per municipality"), linewidth = 0.8) +
  
  geom_line(
    aes(
      y = wolfPopSize_correcteddata / coeff_income,
      linetype = "Wolf population size (corrected)"
    ),
    colour = "black",
    linewidth = 0.8
  ) +
  
  geom_line(
    aes(
      y = wolfPopSize_NOTcorrecteddata_from_annual_reports / coeff_income,
      linetype = "Wolf population size (from reports)"
    ),
    colour = "darkgrey",
    linewidth = 0.8
  ) +
  
  scale_linetype_manual(
    name = "",
    values = c("Wolf population size (corrected)" = 1,
               "Wolf population size (from reports)" = 2)
  ) +
  
  scale_color_manual(
    name = "",
    values = c("Average annual income per municipality" = "#006600")
  ) +
  
  scale_y_continuous(
    name = "Individual income",
    sec.axis = sec_axis(~ . * coeff_income, name = "Wolf population size", breaks = seq(0, 80, by = 10))
  ) +
  
  theme_light() +
  theme(
    axis.title.y = element_text(size = 18),
    axis.title.y.right = element_text(size = 18),
    axis.title.x = element_text(size = 18),
    legend.position = "bottom",
    legend.box.margin = margin(-12, 0, -5, 0),
    legend.text = element_text(size = 18),
    axis.text = element_text(size = 16)
  ) +
  
  labs(x = "Year")


###
plot_bear <- ggplot(df.year_corrected, aes(x = year)) +
  
  geom_line(aes(y = bear_density_live, colour = "Bear harvest (number/km2)"), linewidth = 0.8) +
  
  geom_line(
    aes(
      y = wolfPopSize_correcteddata / coeff_bear,
      linetype = "Wolf population size (corrected)"
    ),
    colour = "black",
    linewidth = 0.8
  ) +
  
  geom_line(
    aes(
      y = wolfPopSize_NOTcorrecteddata_from_annual_reports / coeff_bear,
      linetype = "Wolf population size (from reports)"
    ),
    colour = "darkgrey",
    linewidth = 0.8
  ) +
  
  scale_linetype_manual(
    name = "",
    values = c("Wolf population size (corrected)" = 1,
               "Wolf population size (from reports)" = 2)
  ) +
  
  scale_color_manual(
    name = "",
    values = c("Bear harvest (number/km2)" = "#0066CC")
  ) +
  
  scale_y_continuous(
    name = "Bear harvest (number/km2)",
    sec.axis = sec_axis(~ . * coeff_bear, name = "Wolf population size", breaks = seq(0, 80, by = 10))
  ) +
  
  theme_light() +
  theme(
    axis.title.y = element_text(size = 18),
    axis.title.y.right = element_text(size = 18),
    axis.title.x = element_text(size = 18),
    legend.position = "bottom",
    legend.box.margin = margin(-12, 0, -5, 0),
    legend.text = element_text(size = 18),
    axis.text = element_text(size = 16)
  ) +
  
  labs(x = "Year")


###
plot_moose <- ggplot(df.year_corrected, aes(x = year)) +
  
  geom_line(aes(y = moose_harvest_density, colour = "Moose harvest (number/1000ha)"), linewidth = 0.8) +
  
  geom_line(
    aes(
      y = wolfPopSize_correcteddata / coeff_moose,
      linetype = "Wolf population size (corrected)"
    ),
    colour = "black",
    linewidth = 0.8
  ) +
  
  geom_line(
    aes(
      y = wolfPopSize_NOTcorrecteddata_from_annual_reports / coeff_moose,
      linetype = "Wolf population size (from reports)"
    ),
    colour = "darkgrey",
    linewidth = 0.8
  ) +
  
  scale_linetype_manual(
    name = "",
    values = c("Wolf population size (corrected)" = 1,
               "Wolf population size (from reports)" = 2)
  ) +
  
  scale_color_manual(
    name = "",
    values = c("Moose harvest (number/1000ha)" = "#FF9999")
  ) +
  
  scale_y_continuous(
    name = "Moose harvest",
    sec.axis = sec_axis(~ . * coeff_moose, name = "Wolf population size", breaks = seq(0, 80, by = 10))
  ) +
  
  theme_light() +
  theme(
    axis.title.y = element_text(size = 18),
    axis.title.y.right = element_text(size = 18),
    axis.title.x = element_text(size = 18),
    legend.position = "bottom",
    legend.box.margin = margin(-12, 0, -5, 0),
    legend.text = element_text(size = 18),
    axis.text = element_text(size = 16)
  ) +
  
  labs(x = "Year")


###
plot_wolfNeigh <- ggplot(df.year_corrected, aes(x = year)) +
  
  geom_line(aes(y = number_neighbour_terr, colour = "Neighboring wolf territories"), linewidth = 0.8) +
  
  geom_line(
    aes(
      y = wolfPopSize_correcteddata / coeff_wolfNeigh,
      linetype = "Wolf population size (corrected)"
    ),
    colour = "black",
    linewidth = 0.8
  ) +
  
  geom_line(
    aes(
      y = wolfPopSize_NOTcorrecteddata_from_annual_reports / coeff_wolfNeigh,
      linetype = "Wolf population size (from reports)"
    ),
    colour = "darkgrey",
    linewidth = 0.8
  ) +
  
  scale_linetype_manual(
    name = "",
    values = c("Wolf population size (corrected)" = 1,
               "Wolf population size (from reports)" = 2)
  ) +
  
  scale_color_manual(
    name = "",
    values = c("Neighboring wolf territories" = "#9966CC")
  ) +
  
  scale_y_continuous(
    name = "Number of neighboring wolf territories",
    sec.axis = sec_axis(~ . * coeff_wolfNeigh, name = "Wolf population size", breaks = seq(0, 80, by = 10))
  ) +
  
  theme_light() +
  theme(
    axis.title.y = element_text(size = 18),
    axis.title.y.right = element_text(size = 18),
    axis.title.x = element_text(size = 18),
    legend.position = "bottom",
    legend.box.margin = margin(-12, 0, -5, 0),
    legend.text = element_text(size = 18),
    axis.text = element_text(size = 16)
  ) +
  
  labs(x = "Year")


###
plot_terrRug <- ggplot(df.year_corrected, aes(x = year)) +
  
  # Ruggedness line (colored)
  geom_line(aes(y = ruggedness, colour = "Terrain ruggedness"), linewidth = 0.8) +
  
  # Corrected wolf population (black solid line)
  geom_line(
    aes(
      y = wolfPopSize_correcteddata / coeff_terrRug,
      linetype = "Wolf population size (corrected)"
    ),
    colour = "black",
    linewidth = 0.8
  ) +
  
  # Uncorrected wolf population (grey dashed line)
  geom_line(
    aes(
      y = wolfPopSize_NOTcorrecteddata_from_annual_reports / coeff_terrRug,
      linetype = "Wolf population size (from reports)"
    ),
    colour = "darkgrey",
    linewidth = 0.8
  ) +
  
  # Manually define linetypes and labels
  scale_linetype_manual(
    name = "",
    values = c("Wolf population size (corrected)" = 1,
               "Wolf population size (from reports)" = 2),
    labels = c("Wolf population size (corrected)",
               "Wolf population size (from reports)")
  ) +
  
  # Manual color scale for ruggedness line
  scale_color_manual(
    name = "",
    values = c("Terrain ruggedness" = "brown")
  ) +
  
  scale_y_continuous(
    name = "Terrain ruggedness",
    sec.axis = sec_axis(
      ~ . * coeff_terrRug,
      name = "Wolf population size",
      breaks = seq(0, 80, by = 10)
    )
  ) +
  
  theme_light() +
  theme(
    axis.title.y = element_text(size = 18),
    axis.title.y.right = element_text(size = 18),
    axis.title.x = element_text(size = 18),
    legend.position = "bottom",
    legend.box.margin = margin(-12, 0, -5, 0),
    legend.text = element_text(size = 18),
    axis.text = element_text(size = 16)
  ) +
  
  labs(x = "Year")
```


```{r fig.width=30, fig.height=15, dpi=300}
plots <- list(plot_wolvesKilled, plot_dogs, plot_sheep, plot_terrRug, plot_income, plot_bear, plot_moose, plot_wolfNeigh)
plots <- lapply(plots, function(p) p + theme(text = element_text(size = 2)))

plot_grid(plot_wolvesKilled,plot_dogs, plot_sheep, plot_moose, ncol = 2)

plot_grid(plot_bear, plot_wolfNeigh, plot_income, plot_terrRug, ncol = 2)
```

```{r fig.width=8, fig.height=8, dpi=300}
# Combine the plots into a grid layout
combined_plot <- (plot_dogs | plot_sheep) /
  (plot_wolvesKilled | plot_wolfNeigh) /
  (plot_terrRug | plot_income) /
  (plot_bear  | plot_moose)

# Display the grid
combined_plot + plot_layout(guides = "collect") & theme(legend.position = "bottom")

combined_plot + 
  plot_layout(guides = "collect") & 
  theme(
    legend.position = "bottom",
    text = element_text(size = 10),              # Base text size
    axis.title = element_text(size = 10),        # Axis titles
    axis.text = element_text(size = 8),          # Axis tick labels
    legend.text = element_text(size = 9)         # Legend text
  )

```



## 4.3. Municipality level

```{r}
# summarise territory data per municipality
# slice_max is used to select only one row per territory per year
# all variables are the same for both individuals except for inbreeding
# however, with inbreeding_disapp we also have the same inbreeding value for both individuals; thus, the individual chosen is irrelevant and random

municipalityTerritoryStats <-
  dataToModel7 %>%
  group_by(territory, autumn) %>%
  slice_max(inbreeding_disapp, with_ties = FALSE) %>% #returns the first row when value is the same
  ungroup() %>%
  arrange(territory, autumn) %>%
  dplyr::mutate(year = autumn) %>%
  group_by(NAME_2) %>%
  dplyr::summarise(
    wolfPopSize = length(unique(territory)),
    disappearance = sum(fate == "disapp", na.rm = TRUE),
    n_territories_without_disap = wolfPopSize - disappearance,
    prop_disappearance = disappearance / wolfPopSize,
  ) 

#head(municipalityTerritoryStats, n=30)
# proportion of disappearance per municipality
p <-
  ggplot(municipalityTerritoryStats,
         aes(x = reorder(NAME_2,+prop_disappearance),
             y = prop_disappearance)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90))
p

municipalityTerritoryFeatures <-
  territoriesToModel2 %>%
  dplyr::mutate(year = autumn) %>%
  #dplyr::filter(Fate_binary == 1) %>% # filter only disappearances
  dplyr::group_by(NAME_2) %>%
  dplyr::summarise(
    ruggedness = mean(ruggedness, na.rm = TRUE),
    gravel = mean(gravel, na.rm = TRUE),
    paved = mean(paved, na.rm = TRUE),
    human_density = mean(human_density, na.rm = TRUE),
    snow_cover = mean(snow_cover, na.rm = TRUE),
    night_light = mean(night_light, na.rm = TRUE),
    elevation_mean = mean(elevation_mean, na.rm = TRUE),
    elevation_stdv = mean(elevation_stdv, na.rm = TRUE),
    moose_harvest_density = mean(moose_harvest_density, na.rm = TRUE),
    mean_inbreeding = mean(mean_inbreeding, na.rm = TRUE),
    max_inbreeding = mean(max_inbreeding, na.rm = TRUE),
    inbreeding_disapp = mean(inbreeding_disapp, na.rm = TRUE),
    number_neighbour_terr = mean(number_neighbour_terr, na.rm = TRUE),
    bear_density_shot = mean(bear_density_shot, na.rm = TRUE),
    bear_density_live = mean(bear_density_live, na.rm = TRUE),
    wolfLKill_n = sum(wolfLKill_n, na.rm = TRUE),
    wolfLKill_cumall = sum(wolfLKill_last5y, na.rm = TRUE),
    wolfLKill_last5y = sum(wolfLKill_last5y, na.rm = TRUE),
    sheepAttacks_n = sum(sheepAttacks_n, na.rm = TRUE),
    NoAffectedSheep_last5y = sum(NoAffectedSheep_last5y, na.rm = TRUE),
    NoAffectedSheep_cumall = sum(NoAffectedSheep_cumall, na.rm = TRUE),
    NoAffectedSheep = sum(NoAffectedSheep, na.rm = TRUE),
    AttackDogs_n = sum(AttackDogs_n, na.rm = TRUE),
    NoAffectedDogs_last5y = sum(NoAffectedDogs_last5y, na.rm = TRUE),
    NoAffectedDogs_cumall = sum(NoAffectedDogs_cumall, na.rm = TRUE),
    NoAffectedDogs = mean(NoAffectedDogs, na.rm = TRUE),
    income = mean(income, na.rm = TRUE),
    n = n()
  ) %>%
  arrange(NAME_2)

municipalityTerritoryStats <-
  municipalityTerritoryStats %>%
  dplyr::filter(
    NAME_2 %in% unique(municipalityTerritoryFeatures$NAME_2)
  )

df.municipality <-
  left_join(
    municipalityTerritoryStats,
    municipalityTerritoryFeatures,
    by = c('NAME_2' = 'NAME_2')
  )

# scale covariates
df.municipality$ruggedness_sc <- scale(df.municipality$ruggedness)[,1]
df.municipality$gravel_sc <- scale(df.municipality$gravel)[,1]
df.municipality$paved_sc <- scale(df.municipality$paved)[,1]
df.municipality$human_density_sc <- scale(df.municipality$human_density)[,1]               
df.municipality$snow_cover_sc <- scale(df.municipality$snow_cover)[,1]
df.municipality$moose_harvest_density_sc <- scale(df.municipality$moose_harvest_density)[,1]  
df.municipality$number_neighbour_terr_sc <- scale(df.municipality$number_neighbour_terr)[,1]
df.municipality$bear_density_shot_sc <- scale(df.municipality$bear_density_shot)[,1]
df.municipality$bear_density_live_sc <- scale(df.municipality$bear_density_live)[,1]
df.municipality$night_light_sc <- scale(df.municipality$night_light)[,1]
df.municipality$elevation_mean_sc <- scale(df.municipality$elevation_mean)[,1]
df.municipality$elevation_stdv_sc <- scale(df.municipality$elevation_stdv)[,1]
df.municipality$inbreeding_sc <- scale(df.municipality$inbreeding_disapp)[,1] 
df.municipality$mean_inbreeding_sc <- scale(df.municipality$mean_inbreeding)[,1] 
df.municipality$max_inbreeding_sc <- scale(df.municipality$max_inbreeding)[,1]
df.municipality$inbreeding_disapp_sc <- scale(df.municipality$inbreeding_disapp)[,1]
df.municipality$wolfLKill_cumall_sc <- scale(df.municipality$wolfLKill_cumall)[,1]  
df.municipality$wolfLKill_last5y_sc <- scale(df.municipality$wolfLKill_last5y)[,1]     
df.municipality$wolfLKill_n_sc <- scale(df.municipality$wolfLKill_n)[,1]  
df.municipality$NoAffectedSheep_cumall_sc <- scale(df.municipality$NoAffectedSheep_cumall)[,1]
df.municipality$NoAffectedSheep_last5y_sc <- scale(df.municipality$NoAffectedSheep_last5y)[,1]
df.municipality$NoAffectedSheep_sc <- scale(df.municipality$NoAffectedSheep)[,1]        
df.municipality$sheepAttacks_n_sc <- scale(df.municipality$sheepAttacks_n)[,1]  
df.municipality$NoAffectedDogs_cumall_sc <- scale(df.municipality$NoAffectedDogs_cumall)[,1]  
df.municipality$NoAffectedDogs_last5y_sc <- scale(df.municipality$NoAffectedDogs_last5y)[,1]
df.municipality$NoAffectedDogs_sc <- scale(df.municipality$NoAffectedDogs)[,1]
df.municipality$AttackDogs_n_sc <- scale(df.municipality$AttackDogs_n)[,1]          
df.municipality$income_sc <- scale(df.municipality$income)[,1] 
df.municipality$wolfPopSize_sc <- scale(df.municipality$wolfPopSize)[,1]

write.csv2(df.municipality,
           'modelInputData_new/df.municipality.csv')

writexl::write_xlsx(df.municipality,"modelInputData_new/df.municipality_new.xlsx")

```

### 4.3.1. Correlations 

```{r}
# Note: covariate correlation had to be calculated for the different scales as big discrepancies between scales were detected
colnames(df.municipality)
covToCorrelate_mun <- df.municipality[c("wolfPopSize","ruggedness","paved","gravel","snow_cover","human_density","moose_harvest_density","bear_density_shot","bear_density_live","night_light","elevation_mean","elevation_stdv","number_neighbour_terr","mean_inbreeding","max_inbreeding","inbreeding_disapp","wolfLKill_cumall","wolfLKill_last5y","wolfLKill_n","NoAffectedSheep_cumall","NoAffectedSheep_last5y","NoAffectedSheep","sheepAttacks_n","NoAffectedDogs_cumall","NoAffectedDogs_last5y","NoAffectedDogs","AttackDogs_n","income")]

cormat <- round(
  cor(
    covToCorrelate_mun,
    use = "pairwise.complete.obs",
    method = "spearman"),
  2)

cormat
upper_tri <- get_upper_tri(cormat)
melted_cormat <- reshape2::melt(upper_tri, na.rm = TRUE)

p1_mun <- ggplot(melted_cormat, aes(Var2, Var1, fill = value))+
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  ggtitle(label = "Spearman correlation - all covariates") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1,
                                   hjust = 1, size = 8)) +
  theme(axis.text.y = element_text(size = 8)) +
  geom_text(aes(Var2, Var1, label = value), color = "white", size = 3)
p1_mun  

# highlyCorrelatedVars <- melted_cormat %>%
#   filter(value >0.5 | value < -0.5)
# #highlyCorrelatedVars

## subset the covariates to keep
subset_keep <- df.municipality[c("wolfPopSize","ruggedness","paved","gravel","snow_cover","human_density","moose_harvest_density","bear_density_live","night_light","elevation_stdv","number_neighbour_terr","inbreeding_disapp","wolfLKill_last5y","NoAffectedSheep_last5y","NoAffectedDogs_last5y","income")]

# change covariates names
colnames(subset_keep) <- c("WolfPopSize","TerrRug","PavedRoad","GravelRoad",
                           "SnowCover","HumanDens","MooseHarv","BearDens",
                           "NightLight","ElevationSTDV","WolfNeighbourTerr",
                           "Inbreeding","WolfHarv_last5y","SheepAttacks_last5y",
                           "DogAttacks_last5y","Income")


## correlation between covariates representing the same mechanism --------------

VarsToTest2 <- c("TerrRug",
                "PavedRoad",
                "GravelRoad",
                "SnowCover",
                "HumanDens",
                "NightLight",
                "ElevationSTDV")

VarsToTest2 <- c("DogAttacks_last5y",
                "SheepAttacks_last5y",
                "WolfHarv_last5y",
                "MooseHarv",
                "Income")  

VarsToTest2 <- c("BearDens",
                 "WolfNeighbourTerr")

sub_VarsToTest2 <- subset_keep %>%
  select(all_of(VarsToTest2))

sub_cormat2 <- round(
  cor(
    sub_VarsToTest2,
    use = "pairwise.complete.obs",
    method = "spearman"),2
)

sub_upper_tri2 <- get_upper_tri(sub_cormat2)
sub_melted_cormat2 <-
  reshape2::melt(sub_upper_tri2, na.rm = TRUE)

p3_sub <- ggplot(sub_melted_cormat2, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  ggtitle(label = "Spearman correlation - Municipality level - Competition") +
  #change title depending on the hypothesis
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1,
    size = 16
  )) +
  theme(axis.text.y = element_text(size = 16)) +
  geom_text(aes(Var2, Var1, label = value),
            color = "white",
            size = 6)
p3_sub

## correlations between chosen variables ---------------------------------------

sub_cormat3 <- round(
  cor(
    subset_keep,
    use = "pairwise.complete.obs",
    method = "spearman"),2
)

sub_upper_tri3 <- get_upper_tri(sub_cormat3)
sub_melted_cormat3 <-
  reshape2::melt(sub_upper_tri3, na.rm = TRUE)

p4_sub <- ggplot(sub_melted_cormat3, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  ggtitle(label = "Spearman correlation - Municipality level") +
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1,
    size = 16
  )) +
  theme(axis.text.y = element_text(size = 14)) +
  geom_text(aes(Var2, Var1, label = value),
            color = "white",
            size = 5)
p4_sub


## correlations of final covariates --------------------------------------------

subset_keep2 <- subset_keep[c("WolfPopSize","TerrRug","MooseHarv",
                              "IncomeProp","WolfNeighbourTerr","Mean_InbreedCoef")]

sub_cormat4 <- round(
  cor(
    subset_keep2,
    use = "pairwise.complete.obs",
    method = "spearman"),2
)

sub_upper_tri4 <- get_upper_tri(sub_cormat4)
sub_melted_cormat4 <-
  reshape2::melt(sub_upper_tri4, na.rm = TRUE)

p5_sub <- ggplot(sub_melted_cormat4, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  ggtitle(label = "Spearman correlation - Municipality level - Final variables") +
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1,
    size = 16
  )) +
  theme(axis.text.y = element_text(size = 16)) +
  geom_text(aes(Var2, Var1, label = value),
            color = "white",
            size = 6)
p5_sub
```

### 4.3.2. Calculate min, max and mean

```{r}

min_cov <- format(sapply(df.municipality,min, na.rm = TRUE), scientific = FALSE)
min_cov <- as.data.frame(min_cov)

max_cov <- format(sapply(df.municipality,max, na.rm = TRUE), scientific = FALSE)
max_cov <- as.data.frame(max_cov)

mean_cov <- format(sapply(df.municipality,mean, na.rm = TRUE), scientific = FALSE)
mean_cov <- as.data.frame(mean_cov)

var <- data.frame(min_cov, max_cov, mean_cov)

var <- var[c("wolfPopSize","ruggedness","paved","gravel","human_density","snow_cover","night_light","NoAffectedSheep_last5y","NoAffectedDogs_last5y","moose_harvest_density","income","wolfLKill_last5y","inbreeding_disapp","number_neighbour_terr","bear_density_live"),]

write.csv(var, "ModelInputData_new/info_variables_municipality.csv")

```


