---
title: "modelInputData"
author: "Leticia Duarte"
output: 
  html_notebook:
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: inline
---

<style type="text/css">

body{ /* Normal  */
      font-size: 13px;
  }
td {  /* Table  */
  font-size: 12px;
}
h1.title {
  font-size: 20px;
  font-weight: bold;
}
h1 { /* Header 1 */
  font-size: 20px;
  font-weight: bold;
}
h2 { /* Header 2 */
    font-size: 17px;
  font-weight: bold;
}
h3 { /* Header 3 */
    font-size: 15px;
  font-weight: bold;
}
code.r{ /* Code block */
    font-size: 12px;
}
pre{ /* Code block - determines code spacing between lines */
    font-size: 12px;
}
h1, .h1, h2, .h2, h3, .h3, h4, .h4 {
    margin-top: 24xpx;
}
</style>

# 1. Packages

```{r}
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(magrittr)
library(sf)
library(ggplot2)
library(cowplot)
library(remotes)
#remotes::install_github("Tong-Chen/YSX")
library("YSX")
```

# 2. Raw data

```{r}
# import data
data <-
  readr::read_csv2(
    "rawData/FateWolvesTerritory1998_2020_Scandinavia_Final_CDB.csv",
    locale = locale(encoding = "ISO-8859-1")
  )

# data clean
idvars <- c(colnames(data)[1:13], "KommunerNamn", "SLU.ID")
explanatoryvars <- c("ruggedness_mean",
                     "average_gravel_km",
                     "average_paved_km",
                     "artificial_area_total",
                     "pop_mean",
                     "mean_snow",
                     "hunt_afo",
                     "hunt_county2",
                     "Fi",
                     "number_neighbour_terr",
                     "bear_density_BZtiff")

varstoselect <- c(idvars, explanatoryvars)

cleandata <- data %>%
  dplyr::select(all_of(varstoselect)) %>%
  dplyr::mutate(mod_id = 1:nrow(data)) %>% 
  relocate(mod_id, .before = id) %>% 
  dplyr::mutate(Fate = str_replace(Fate, "Legal", "legal")) %>%
  dplyr::mutate(Fate = str_replace(Fate, "Illegal", "illegal")) %>%
  dplyr::mutate(Fate = str_replace(Fate, "censor", "censored")) %>%
  dplyr::mutate(Fate = str_replace(Fate, "cernsored", "censored")) %>%
  dplyr::mutate(Fate = str_replace(Fate, "censoreded", "censored")) %>%
  dplyr::mutate(KommunerNamn = tolower(KommunerNamn)) %>%
  ungroup()
```

# 3. Adding variables

## 3.1. Inbreeding

```{r}
# calculate average and maximum inbreeding per territory
cleandata_withFi <- 
  cleandata %>%
  group_by(territory) %>%
  dplyr::mutate(mean_fi = mean(Fi),
         max_fi = max(Fi))
```


## 3.2. Spatial data

```{r}
norway <-
  sf::st_read("spatialData/gadm36_NOR_shp/gadm36_NOR_2.shp")
sweden <-
  sf::st_read("spatialData/gadm36_SWE_shp/gadm36_SWE_2.shp")
scandinavia <- rbind(norway, sweden) %>%
  select(NAME_1, NAME_2, geometry)

# add municipality (gadm classification)
# convert to spatial data and add gadm names
dataToModel1 <-
  st_as_sf(
    x = cleandata_withFi,
    #convert to spatial features (sf)
    coords = c("X_coordinate_fatefile",
               "Y_coordinate_fatefile"),
    crs = "EPSG:3021"
  ) %>% #RT90
  st_transform(crs = st_crs(scandinavia)) %>%
  st_intersection(scandinavia) %>% # add gadm names for municipalities
  st_drop_geometry() # back to data frame

```

## 3.3. Legally killed wolves

```{r}
# number of legally killed wolves
killedWolvesData <- readr::read_csv("rawData/Rovbase230311_CW_killed wolves_final.csv") %>%
  separate(Dødsdato, c("DødsYear", "DødsMonth","DødsDay")) %>%
  dplyr::mutate(DødsYear = as.numeric(DødsYear),
         DødsMonth = as.numeric(DødsMonth),
         DødsDay = as.numeric(DødsDay)) %>%
  select("RovbaseID", "DødsYear",
         "Nord (RT90)", "Øst (RT90)") %>% # columns of interest 
  dplyr::filter(DødsYear > 1998) # remove records outside of the monitoring period

# convert to spatial data
killedWolvesSpatial <- st_as_sf(x = killedWolvesData,
                                coords = c("Øst (RT90)", "Nord (RT90)"),
                                crs = "EPSG:3021") %>% #RT90
  st_transform(crs = st_crs(scandinavia)) #%>% 
  #select(RovbaseID,geometry)

# intersect legally killed wolves with scandinavia admn areas
wolfLKill <- st_intersection(killedWolvesSpatial, scandinavia)

# count legal kills per municipality per year
wolfLKill_municipality <- wolfLKill %>% dplyr::count(DødsYear, NAME_2) %>%
  dplyr::rename(wolfLKill_n = n)


# build base dataframe 
kommun <- rep(levels(as.factor(dataToModel1$NAME_2)), # uses the kommunes with wolf territories
              each=length(c(2000:2022)))
# data starts in 2000 - see wolfLKill_municipality
year <- rep(c(2000:2022), length(levels(as.factor(dataToModel1$NAME_2))))
df <- data.frame(kommun = kommun,year = year)


# cumulative sum in the last five years include the target year and the previous 4
# cumulative slide was calculated based on https://stackoverflow.com/questions/72587840/how-to-get-the-cummean-of-the-last-n-rows-before-the-last-row-using-dplyr
wolfLKillAllMetrics <- left_join(df,
                                 wolfLKill_municipality,
                  by=c('kommun'='NAME_2', 'year'='DødsYear')) %>%
  #ungroup() %>%
  group_by(kommun) %>%
  # replaces NAs by zeros. This considers that records of wolf kills were recorded since 2000 (first year with records)
  dplyr::mutate_at(c('wolfLKill_n'), ~tidyr::replace_na(.,0)) %>%
  dplyr::mutate(wolfLKill_last5y = slider::slide_dbl(wolfLKill_n, sum, .before = 4, .after = 0),
         .after=year) %>%
  dplyr::mutate(wolfLKill_cumall = cumsum(wolfLKill_n),.after=year)


# join to dkommun# join to data to model  
dataToModel2 <- left_join(dataToModel1,
                          wolfLKillAllMetrics,
                          by=c('NAME_2'='kommun', 'Autumn'='year')) %>%
  # replaces zeros to NAs before 1999 because there was no wolf kill data before 2000
  dplyr::mutate(wolfLKill_n = ifelse(Autumn <= 1999, NA, wolfLKill_n))

```

## 3.4. Wolf attacks on sheep

```{r}
# old raw file
#rovbasedata <- readr::read_csv("data/rawData/rovbaseSample.csv") %>%
  # sums killed, injured, or missing individuals across the different time spans measured
  #dplyr::mutate(NoAffectedSheep = rowSums(across("Antal döda under 1 år":"Antal saknade över 2 år"), na.rm = F))


# new raw files
sheep_data <- readr::read_csv("rawData/Wolf attacks on sheep_Rovbase 230131_Sweden_Norway.csv") %>%
  dplyr::mutate(NoAffectedSheep = Antall)

# select attack id, date and coordinates (check if coordinate system is the same as in scandinavia)
sheep_data2 <- sheep_data[c("Skada på", "Fynddatum",
                                          "NoAffectedSheep", "Kommune",
                                          "Nord (RT90)", "Öst (RT90)")]

sheep_data2$coord <- paste0(sheep_data2$`Nord (RT90)`, ",", sheep_data2$`Öst (RT90)`)

# group rows that represent the same event from norway according to date and location
grouped_sheep <- sheep_data2 %>%
  group_by(Fynddatum,coord) %>% # group by unique date and coordinates
  summarize(
    Kommune = first(Kommune), # keep the first
    `Skada på` = first(`Skada på`),
    `Nord (RT90)` = first(`Nord (RT90)`),
    `Öst (RT90)` = first(`Öst (RT90)`),
    NoAffectedSheep = sum(NoAffectedSheep)  # sum the number of attacks
  ) %>%
  ungroup()  # ungroup to return to a normal data frame

# select attack id, date and coordinates (check if coordinate system is the same as in scandinavia)
killedsheepclean <- grouped_sheep %>% select("Skada på", "Fynddatum",
                                           "NoAffectedSheep", "Kommune",
                                           "Nord (RT90)", "Öst (RT90)") %>%
  na.omit() %>%
  dplyr::filter(`Skada på` %in% c("Får")) %>% # select only sheep
  separate(Fynddatum, c("day","month","year")) %>% 
  dplyr::mutate(year = as.numeric(year)) #%>%
  #filter(year>=1998) 

# read as spatial points and change coordinate to RT90
killedsheepSpatial <- st_as_sf(x = killedsheepclean,
                               coords = c("Öst (RT90)", "Nord (RT90)"),
                               crs = "EPSG:3021") %>% #RT90
  st_transform(crs = st_crs(scandinavia)) %>% 
  select(year, "NoAffectedSheep", geometry) 

# intersect sheep attacks with scandinavia admn areas
sheepAttacks <- st_intersection(killedsheepSpatial, scandinavia)

# count number of sheep killed  and number of attacks per municipality
sheepAffected_municipality <- sheepAttacks %>%
  dplyr::group_by(year, NAME_2) %>%
  dplyr::summarise(NoAffectedSheep = sum(NoAffectedSheep),
            n = n(),.groups = "drop") %>%
  dplyr::rename(sheepAttacks_n = n)

# build base dataframe 
kommun <- rep(levels(as.factor(dataToModel2$NAME_2)), # uses the kommunes with wolf territories
              each=length(c(1995:2021)))
# data starts in 1995 - see sheepAffected_municipality
year <- rep(c(1995:2021), length(levels(as.factor(dataToModel2$NAME_2))))
df <- data.frame(kommun = kommun,year = year)

# cumulative metrics
# cumulative sum in the last five years include the target year and the previous 4
# cumulative slide was calculated based on https://stackoverflow.com/questions/72587840/how-to-get-the-cummean-of-the-last-n-rows-before-the-last-row-using-dplyr
sheepAllMetrics <- left_join(df,
                             sheepAffected_municipality,
                  by=c('kommun'='NAME_2', 'year'='year')) %>%
  dplyr::group_by(kommun) %>%
  # replaces NAs by zeros. This considers that records of sheep events were recorded since 1995 (first year with records)
  mutate_at(c('NoAffectedSheep','sheepAttacks_n'), ~replace_na(.,0)) %>%
  dplyr::mutate(NoAffectedSheep_last5y = slider::slide_dbl(NoAffectedSheep, sum, .before = 4, .after = 0),
         .after=year) %>%
  dplyr::mutate(NoAffectedSheep_cumall = cumsum(NoAffectedSheep),.after=year)
 
# join to data to model  
dataToModel3 <- left_join(dataToModel2,
                      sheepAllMetrics,
                      by=c('NAME_2'='kommun', 'Autumn'='year'))

```

### 3.4.1. Add variable of all wolf attacks on sheep (includes outside of municipalities with territories)

```{r}

# build base dataframe 
kommun <- rep(levels(as.factor(sheepAffected_municipality$NAME_2)), # uses all kommunes (including without wolf territories)
              each=length(c(1995:2021)))
# data starts in 1995 - see sheepAffected_municipality
year <- rep(c(1995:2021), length(levels(as.factor(sheepAffected_municipality$NAME_2))))
df <- data.frame(kommun = kommun,year = year)

# cumulative metrics
# cumulative sum in the last five years include the target year and the previous 4
# cumulative slide was calculated based on https://stackoverflow.com/questions/72587840/how-to-get-the-cummean-of-the-last-n-rows-before-the-last-row-using-dplyr
sheepAllMetrics2 <- left_join(df,
                             sheepAffected_municipality,
                  by=c('kommun'='NAME_2', 'year'='year')) %>%
  dplyr::group_by(kommun) %>%
  # replaces NAs by zeros. This considers that records of sheep events were recorded since 1995 (first year with records)
  mutate_at(c('NoAffectedSheep','sheepAttacks_n'), ~replace_na(.,0)) %>%
  dplyr::mutate(NoAffectedSheep_last5y_v2 = slider::slide_dbl(NoAffectedSheep, sum, .before = 4, .after = 0),
         .after=year) %>%
  dplyr::mutate(NoAffectedSheep_cumall_v2 = cumsum(NoAffectedSheep),.after=year)
 
# join to data to model  
dataToModel3_v2 <- left_join(sheepAllMetrics2,
                            dataToModel3,
                            by=c('kommun'='NAME_2', 'year'='Autumn'))


```

## 3.5. Wolf attacks on dogs

```{r}
dogData <- readr::read_csv("rawData/Wolf attacks on dogs Sweden and Norway_Rovbase 230131_230210_CW.csv") %>%
  # sums killed, injured, or missing individuals across the different time spans measured
  #dplyr::mutate(NoAffectedDogs = rowSums(across("Antal döda under 1 år":"Antal saknade över 2 år"), na.rm = F)) #excludes data from Norway
  dplyr::mutate(NoAffectedDogs = Antall) #includes data from Norway

# select attack id, date and coordinates (check if coordinate system is the same as in scandinavia)
AttacksDogclean <- dogData %>% select("HändelseID", "Country", "Fynddatum", "NoAffectedDogs", "Kommune",
                                   "Nord (RT90)", "Öst (RT90)") %>%
  na.omit() %>%
  separate(Fynddatum, c("year","month","day")) %>%
  dplyr::filter(year >= 1995) %>% # original span is 1995-2022
  dplyr::mutate(year = as.numeric(year))

# read as spatial points and change coordinate to RT90
AttacksDogSpatial <- st_as_sf(x = AttacksDogclean,
                           coords = c("Öst (RT90)", "Nord (RT90)"),
                           crs = "EPSG:3021") %>% #RT90
  st_transform(crs = st_crs(scandinavia)) %>% 
  select(HändelseID, NoAffectedDogs, year, geometry)

# intersect attacks on dogs with scandinavia admn areas
AttackDogs <- st_intersection(AttacksDogSpatial, scandinavia)

# count number of attacks on dogs per year and per municipality
AttackDogs_municipality <- AttackDogs %>%
  group_by(year, NAME_2) %>%
  dplyr::summarise(n = n(),.groups = "drop") %>%
  dplyr::rename(AttackDogs_n = n)

# build base dataframe 
kommun <- rep(levels(as.factor(dataToModel3$NAME_2)),
              each=length(c(1995:2021)))
# data starts in 1995 - see AttackDogs_municipality
year <- rep(c(1995:2021), length(levels(as.factor(dataToModel3$NAME_2))))
df <- data.frame(kommun = kommun,year = year)

# cumulative metrics
# cumulative sum in the last five years include the target year and the previous 4
# cumulative slide was calculated based on https://stackoverflow.com/questions/72587840/how-to-get-the-cummean-of-the-last-n-rows-before-the-last-row-using-dplyr
dogAllMetrics <- left_join(df,
                           AttackDogs_municipality,
                           by=c('kommun'='NAME_2', 'year'='year')) %>%
  dplyr::group_by(kommun) %>%
  # replaces NAs by zeros. This considers that records of sheep events were recorded since 1995
  # (first year with records)
  mutate_at(c('AttackDogs_n'), ~replace_na(.,0)) %>%
  dplyr::mutate(AttackDogs_last5y = slider::slide_dbl(AttackDogs_n, sum, .before = 4, .after = 0),
         .after=year) %>%
  dplyr::mutate(AttackDogs_cumall = cumsum(AttackDogs_n),.after=year)

# join to data to model  
dataToModel4 <- left_join(dataToModel3,
                          dogAllMetrics,
                          by=c('NAME_2'='kommun', 'Autumn'='year'))
```


## 3.6. Individual income

```{r}

# Load necessary library
library(readr)
library(dplyr)
library(readxl)
library(tidyr)

# notes for norway data download

# Rømskog municipality in Norway no longer exists under that name.
# On January 1, 2020, Rømskog merged with Aurskog-Høland municipality
# in Viken County. So now, Rømskog is part of Aurskog-Høland.
# norway average gross individual income was extracted from https://www.ssb.no/en/statbank/table/05854/
# codes in kommun original were removed manually in excel, column names were also edited manually 


# Load the CSV file
file_path <- "data/rawData/income_forWolf/05854_MainEntriesFromTheTaxAssessmentForResidents17yearsAndOlder_individualIncomeNorway_edited.csv"
df <- read_delim(file_path, delim = ";", col_types = cols(.default = "c"))

# Delete the column 'kommun subset'
filtered_df <- df %>% select(-`kommun original`, -`age`, -`metric`)

# Convert all columns, except column "kommun", to numeric
filtered_df <- filtered_df %>%
  mutate(across(-kommun, ~ suppressWarnings(as.numeric(.))))

# Calculate row means and then summarize to one row per group
df_sum <- filtered_df %>%
  group_by(kommun) %>%
  summarize(across(where(is.numeric), sum, na.rm = TRUE))

# Replace zeros with NAs
df_sum <- df_sum %>%
  mutate(across(where(is.numeric), ~ na_if(.x, 0))) %>%
  mutate(country = "Norway") %>%
  select(country, everything())

# View the filtered dataframe
View(df_sum)

# notes for sweden data download
# excel was manually edited to remove first row and last rows
# kommun name only available at the first row for each kommun and was manually inputed for the remaining ages
# Before 2003 Knivsta was part of Uppsala Municipality and hence the NA values 
# swedish income data was extracted from Total earned income for residents in Sweden throughout the year by region, gender, age and income class. Years 1999 - 2023
# https://www.statistikdatabasen.scb.se/pxweb/sv/ssd/START__HE__HE0110__HE0110A/SamForvInk1/
# kommun names for all the age groups were manually edited as the original table only had them for the first line of each kommun 
# needed to work in excel at this step since the kommun names in Swedish were not reading correctly when trying with csv

# Load the Excel file
file_path2 <- "data/rawData/income_forWolf/Sweden_HE0110J7_20250301-210646_edited.xlsx"
df2 <- read_excel(file_path2)
View(df2)
# Remove columns
df2 <- df2 %>% select(-age, -metric)

# Convert all columns, except column "kommun", to numeric
df2 <- df2 %>%
  mutate(across(-kommun, ~ suppressWarnings(as.numeric(.))))

# Calculate row means and then summarize to one row per group
df2_avg <- df2 %>%
  group_by(kommun) %>%
  summarize(across(where(is.numeric), mean))

# Multiply all numeric values by a thousand and reduce decimal places to zero
# This is because the values in the Excel file are in K SEK
df2_avg <- df2_avg %>%
  mutate(across(where(is.numeric), ~ round(.x * 1000, 0))) %>%
  mutate(country = "Sweden") %>%
  select(country, everything())

View(df2_avg)

# notes merging the two dataframes
# I have ignored that norway data is in NOK and swedish data is in SEK

# Merge df2_avg and df_sum by columns with the same name
merged_df <- full_join(df_sum, df2_avg, by = intersect(names(df_sum), names(df2_avg)))

# View the merged dataframe
View(merged_df)

# Calculate the average of each year
yearly_avg <- merged_df %>%
  summarize(across(where(is.numeric), mean, na.rm = TRUE))

# Divide each cell of a given year by the mean of the same year
normalized_df <- merged_df %>%
  mutate(across(where(is.numeric), ~ . / yearly_avg[[cur_column()]]))

# View the normalized dataframe
View(normalized_df)

# Save the merged dataframe to a new CSV file
# write_csv(normalized_df, "data/rawData/income_forWolf/Merged_Data.csv")

# Pivot normalized_df to have columns: country, kommun, year, and value
pivoted_df <- normalized_df %>%
  pivot_longer(cols = -c(country, kommun), names_to = "year", values_to = "value")

# View the pivoted dataframe
View(pivoted_df)

#dataToModel5 <-
#  left_join(
#    dataToModel4,
#    pivoted_df,
# by = c('NAME_2' = 'Kommun', 'Autumn' = 'Year') columns need to be checked here to match new dataToModel4 
#    )
```


## 3.7. Prepare the data

```{r}
# add years since first detection for test with survival analyses
dataToModel6 <- 
  dataToModel5 %>%
  group_by(territory) %>%
  dplyr::mutate(
    first_yeardet = min(Autumn),
    last_yeardet = max(Autumn),
    years_firstdet = Autumn - first_yeardet,
    time_start = Autumn - first_yeardet,
    time_end = Spring - first_yeardet
  ) 

# add population size
# note: in this way legally removed territories in one year or territories that
# disappeared due to natural causes are still considered in the census of that
# spring. The same for other causes of death 
wolfPopSize <-
  dataToModel6 %>%
  group_by(Spring) %>%
  dplyr::summarise(
    Spring = unique(Spring),
    # number of unique territories per year, independently of their fate
    wolfPopSize = length(unique(territory))
  )

dataToModel7 <-
  left_join(
    dataToModel6,
    wolfPopSize,
    by = c('Spring' = 'Spring')
  )
  
# prepare for correlation analyses
dataToModel8 <- 
  dataToModel7 %>%
  relocate(c(NAME_1, NAME_2), .after = Comment) %>%
  select(-c(geometry.x, geometry.y, geometry)) %>%
  #relocate(geometry, .after = Comment) %>%
  relocate(income_prop, .after = IndividualIncome) %>%
  relocate(wolfPopSize, .after = income_prop)
#mutate_at(c("wolfLKill_cumall", "wolfLKill_n", "sheepAttacks_n",
#            "AttackDogs_cumall", "AttackDogs_n"),
#          ~as.numeric(.))

```


# 4. Model input data

## 4.1. Territory level

```{r}
# territory selection 
# Note: function tested by Cecilia and Giorgia
# territorySelection_testFunctionsCeciliaGiorgia.R
# inconsistencies detected for instance when there is combination of fates such
# as "illegal", "other" - 6 territories and "illegal", "legal" - 12 territories 
territoriesToModel <- dataToModel8 %>%
  group_by(territory) %>%
  # only keep territories that have "NA","illegal" or,"censored" in their fate
  # by excluding other fates
  dplyr::filter(!any(Fate %in% c("other","legal","natural","traffic"))) %>% # 291
  # keep the row with the maximum inbreeding, if tied keep the first row
  group_by(territory, Autumn) %>%
  slice_max(mean_fi, with_ties=FALSE) %>% 
  # add column to double-check potential duplicate rows within territories 
  #mutate(Autumn_dup = duplicated(Autumn)) %>%
  ungroup() %>%
  dplyr::mutate(Fate_binary = ifelse(is.na(Fate), 0,1),.after=Fate) %>%
  arrange(territory, Autumn)

length(unique(territoriesToModel$territory)) # 291
nrow(territoriesToModel) # 642 only one individual per territory
unique(territoriesToModel$Fate)

# count disappearances and ongoing territories per year ------------------------
df <- territoriesToModel %>% 
  group_by(Spring) %>%
  dplyr::summarise(
    ongoing = sum(is.na(Fate)==TRUE),
    disappearance = sum(Fate %in% c("illegal" ,"censored")),
    sum = ongoing + disappearance,
    prop_disappearance = round(disappearance/sum, 2),
    n()
  )

sum(df$disappearance) # 193 # double checking
sum(df$ongoing) # 449 # double checking
print(df, n=30)

# plot disappearance rate - it does not match the plot from Liberg et al. 2020
# I cannot calculate prop. disappearance and total number of territories
# like this because I am excluding territories with other fates (it influences the sum)
# what are the implications for the territory analyses?
# the plot
ggplot(data=df, aes(x=Spring, y=prop_disappearance, group=1)) +
  geom_line()+
  geom_point()

# model data scaled ------------------------------------------------------------
# for some reason scale with dplyr and mutate was producing nan
# count data not scaled
modeldatascaled <- territoriesToModel
modeldatascaled$ruggedness_mean_sc <- scale(modeldatascaled$ruggedness_mean)[,1] 
modeldatascaled$average_gravel_km_sc <- scale(modeldatascaled$average_gravel_km)[,1] 
modeldatascaled$average_paved_km_sc <- scale(modeldatascaled$average_paved_km)[,1]       
modeldatascaled$artificial_area_total_sc <- scale(modeldatascaled$artificial_area_total)[,1] 
modeldatascaled$pop_mean_sc <- scale(modeldatascaled$pop_mean)[,1]               
modeldatascaled$mean_snow_sc <- scale(modeldatascaled$mean_snow)[,1] 
modeldatascaled$hunt_afo_sc <- scale(modeldatascaled$hunt_afo)[,1]              
modeldatascaled$hunt_county2_sc <- scale(modeldatascaled$hunt_county2)[,1] 
modeldatascaled$number_neighbour_terr_sc <- scale(modeldatascaled$number_neighbour_terr)[,1]
modeldatascaled$bear_density_BZtiff_sc <- scale(modeldatascaled$bear_density_BZtiff)[,1] 
modeldatascaled$Fi_sc <- scale(modeldatascaled$Fi)[,1] 
modeldatascaled$mean_fi_sc <- scale(modeldatascaled$mean_fi)[,1] 
modeldatascaled$max_fi_sc <- scale(modeldatascaled$max_fi)[,1]               
# modeldatascaled$wolfLKill_cumall_sc <- scale(modeldatascaled$wolfLKill_cumall)[,1]  # count data - not scaled 
modeldatascaled$wolfLKill_last5y_sc <- scale(modeldatascaled$wolfLKill_last5y)[,1]     
# modeldatascaled$wolfLKill_n_sc <- scale(modeldatascaled$wolfLKill_n)[,1]  # count data - not scaled 
# modeldatascaled$NoAffectedSheep_cumall_sc <- scale(modeldatascaled$NoAffectedSheep_cumall)[,1]  # count data - not scaled 
modeldatascaled$NoAffectedSheep_last5y_sc <- scale(modeldatascaled$NoAffectedSheep_last5y)[,1]
# modeldatascaled$NoAffectedSheep_sc <- scale(modeldatascaled$NoAffectedSheep)[,1]  # count data - not scaled       
# modeldatascaled$sheepAttacks_n_sc <- scale(modeldatascaled$sheepAttacks_n)[,1]  # count data - not scaled 
# modeldatascaled$AttackDogs_cumall_sc <- scale(modeldatascaled$AttackDogs_cumall)[,1]  # count data - not scaled      
modeldatascaled$AttackDogs_last5y_sc <- scale(modeldatascaled$AttackDogs_last5y)[,1]
# modeldatascaled$AttackDogs_n_sc <- scale(modeldatascaled$AttackDogs_n)[,1]  # count data - not scaled         
modeldatascaled$IndividualIncome_sc <- scale(modeldatascaled$IndividualIncome)[,1] 
modeldatascaled$income_prop_sc <- scale(modeldatascaled$income_prop)[,1]
modeldatascaled$wolfPopSize_sc <- scale(modeldatascaled$wolfPopSize)[,1]

df.territory <- as_tibble(modeldatascaled)

# territory data
write.csv(df.territory,'modelInputData/df.territory.csv')

```

### 4.1.1. Correlations

```{r}
## correlations with all variables ---------------------------------------------
colnames(territoriesToModel)
covToCorrelate_terr <- territoriesToModel[18:43]

cormat <- round(
  cor(
    covToCorrelate_terr,
    use = "pairwise.complete.obs",
    method = "spearman"),
  2)

cormat
upper_tri <- get_upper_tri(cormat)
melted_cormat <- reshape2::melt(upper_tri, na.rm = TRUE)

p1_terr <- ggplot(melted_cormat, aes(Var2, Var1, fill = value))+
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  ggtitle(label = "Spearman correlation - territory level - all covariates") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1,
                                   hjust = 1, size = 10)) +
  theme(axis.text.y = element_text(size = 10)) +
  geom_text(aes(Var2, Var1, label = value), color = "white", size = 4.5)
p1_terr

# highlyCorrelatedVars <- melted_cormat %>%
#   filter(value >0.5 | value < -0.5)
#highlyCorrelatedVars

## correlation between metrics for the same variable ---------------------------
VarsToTest <- c("AttackDogs_last5y",
                "AttackDogs_n",
                "AttackDogs_cumall") # keep AttackDogs_last5y

VarsToTest <- c("NoAffectedSheep_cumall",
                "NoAffectedSheep_last5y",
                "NoAffectedSheep",
                "sheepAttacks_n") # keep NoAffectedSheep_last5y

VarsToTest <- c("wolfLKill_cumall",
                "wolfLKill_last5y",
                "wolfLKill_n") # keep wolfLKill_last5y

VarsToTest <- c("hunt_afo",
                "hunt_county2") # not correlated but keep hunt_county2 (density)

VarsToTest <- c("average_gravel_km",
                "average_paved_km") # keep average_gravel_km

VarsToTest <- c("IndividualIncome",
                "income_prop") # keep income_prop

VarsToTest <- c("mean_fi",
                "max_fi") # keep mean_fi


subset <- covToCorrelate_terr %>%
  select(all_of(VarsToTest))

sub_cormat <- round(
  cor(
    subset,
    use = "pairwise.complete.obs",
    method = "spearman"),2
)

sub_upper_tri <- get_upper_tri(sub_cormat)
sub_melted_cormat <-
  reshape2::melt(sub_upper_tri, na.rm = TRUE)

p2_sub <- ggplot(sub_melted_cormat, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  ggtitle(label = "Spearman correlation - Territory level") +
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1,
    size = 16
  )) +
  theme(axis.text.y = element_text(size = 16)) +
  geom_text(aes(Var2, Var1, label = value),
            color = "white",
            size = 6)
p2_sub

# subset the covariates to keep
subset_keep <- territoriesToModel[c("wolfPopSize","ruggedness_mean",
                                    "average_gravel_km","artificial_area_total",
                                    "pop_mean","mean_snow","AttackDogs_last5y",
                                    "NoAffectedSheep_last5y","wolfLKill_last5y",
                                    "hunt_county2","income_prop",
                                    "bear_density_BZtiff","number_neighbour_terr",
                                    "mean_fi")]

# change covariates names
colnames(subset_keep) <- c("WolfPopSize","TerrRug","GravelRoad",
                            "ArtificialArea","HumanDens","SnowCover",
                            "DogAttacks_last5y","SheepAttacks_last5y",
                            "LegWolfHarv_last5y","MooseHarv","IncomeProp",
                            "BearHarv","WolfNeighbourTerr","Mean_InbreedCoef")


## correlation between covariates representing the same mechanism --------------
VarsToTest2 <- c("TerrRug",
                "GravelRoad",
                "ArtificialArea",
                "HumanDens",
                "SnowCover")

VarsToTest2 <- c("DogAttacks_last5y",
                "SheepAttacks_last5y",
                "LegWolfHarv_last5y",
                "MooseHarv",
                "IncomeProp")  

VarsToTest2 <- c("BearHarv",
                "WolfNeighbourTerr")

sub_VarsToTest2 <- subset_keep %>%
  select(all_of(VarsToTest2))

sub_cormat2 <- round(
  cor(
    sub_VarsToTest2,
    use = "pairwise.complete.obs",
    method = "spearman"),2
)

sub_upper_tri2 <- get_upper_tri(sub_cormat2)
sub_melted_cormat2 <-
  reshape2::melt(sub_upper_tri2, na.rm = TRUE)

p3_sub <- ggplot(sub_melted_cormat2, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  ggtitle(label = "Spearman correlation - Territory level - Accessibility") +
  #change title depending on the hypothesis
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1,
    size = 16
  )) +
  theme(axis.text.y = element_text(size = 16)) +
  geom_text(aes(Var2, Var1, label = value),
            color = "white",
            size = 6)
p3_sub


## correlations between chosen variables ---------------------------------------

sub_cormat3 <- round(
  cor(
    subset_keep,
    use = "pairwise.complete.obs",
    method = "spearman"),2
)

sub_upper_tri3 <- get_upper_tri(sub_cormat3)
sub_melted_cormat3 <-
  reshape2::melt(sub_upper_tri3, na.rm = TRUE)

p4_sub <- ggplot(sub_melted_cormat3, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  ggtitle(label = "Spearman correlation - Territory level - Chosen variables") +
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1,
    size = 16
  )) +
  theme(axis.text.y = element_text(size = 16)) +
  geom_text(aes(Var2, Var1, label = value),
            color = "white",
            size = 6)
p4_sub



## correlations of final covariates --------------------------------------------

subset_keep2 <- subset_keep[c("WolfPopSize","TerrRug","GravelRoad",
                              "ArtificialArea","HumanDens",
                              "DogAttacks_last5y","SheepAttacks_last5y",
                              "LegWolfHarv_last5y","IncomeProp",
                              "BearHarv","WolfNeighbourTerr","Mean_InbreedCoef")]

sub_cormat4 <- round(
  cor(
    subset_keep2,
    use = "pairwise.complete.obs",
    method = "spearman"),2
)

sub_upper_tri4 <- get_upper_tri(sub_cormat4)
sub_melted_cormat4 <-
  reshape2::melt(sub_upper_tri4, na.rm = TRUE)

p5_sub <- ggplot(sub_melted_cormat4, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  ggtitle(label = "Spearman correlation - Territory level - Final variables") +
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1,
    size = 16
  )) +
  theme(axis.text.y = element_text(size = 16)) +
  geom_text(aes(Var2, Var1, label = value),
            color = "white",
            size = 6)
p5_sub

```

### 4.1.2. Calculate min, max and mean

```{r}
min_cov <- format(sapply(df.territory,min, na.rm = TRUE), scientific = FALSE)
min_cov <- as.data.frame(min_cov)

max_cov <- format(sapply(df.territory,max, na.rm = TRUE), scientific = FALSE)
max_cov <- as.data.frame(max_cov)

mean_cov <- format(sapply(df.territory,mean, na.rm = TRUE), scientific = FALSE)
mean_cov <- as.data.frame(mean_cov)

var <- data.frame(min_cov, max_cov, mean_cov)

write.csv(var, "info_variables_territory.csv")

```

## 4.2. Population level

```{r}
# summarise territory data per year
# adopted from territory selection but includes all fates to calculate pop.size
yearlyTerritoryStats <- 
  dataToModel8 %>%
  group_by(territory, Spring) %>%
  slice_max(mean_fi, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(territory, Spring) %>%
  dplyr::mutate(year = Spring, .before = mod_id) %>%
  group_by(year) %>%
  dplyr::summarise(
    wolfPopSize = length(unique(territory)),
    disappearance = sum(Fate == "illegal", na.rm = TRUE),
    n_territories_without_disap = wolfPopSize - disappearance,
    prop_disappearance = disappearance / wolfPopSize,
  )
#head(yearlyTerritoryStats, n=30)


yearlyTerritoryFeatures <- territoriesToModel %>%
  dplyr::mutate(year = Spring, .before = mod_id) %>%
  group_by(year) %>%
  #dplyr::filter(Fate_binary == 1) %>% # filter only disappearances
  dplyr::summarise(
    mean_ruggedness = mean(ruggedness_mean, na.rm = TRUE),
    average_gravel_km = mean(average_gravel_km, na.rm = TRUE),
    average_paved_km = mean(average_paved_km, na.rm = TRUE),
    pop_mean = mean(pop_mean, na.rm = TRUE),
    mean_snow = mean(mean_snow, na.rm = TRUE),
    mean_artificial_area_total = mean(artificial_area_total, na.rm = TRUE),
    hunt_county2 = mean(hunt_county2, na.rm = TRUE),
    mean_fi = mean(mean_fi, na.rm = TRUE),
    mean_maxfi = mean(max_fi, na.rm = TRUE),
    mean_number_neighbour_terr = mean(number_neighbour_terr, na.rm = TRUE),
    mean_bear_density_BZtiff = mean(bear_density_BZtiff, na.rm = TRUE),
    sum_wolfLKill_n = sum(wolfLKill_n, na.rm = TRUE),
    sum_wolfLKill_last5y = sum(wolfLKill_last5y, na.rm = TRUE),
    sum_sheepAttacks_n = sum(sheepAttacks_n, na.rm = TRUE),
    sum_NoAffectedSheep_last5y = sum(NoAffectedSheep_last5y, na.rm = TRUE),
    sum_AttackDogs_n = sum(AttackDogs_n, na.rm = TRUE),
    sum_AttackDogs_last5y = sum(AttackDogs_last5y, na.rm = TRUE),
    mean_IndividualIncome = mean(IndividualIncome, na.rm = TRUE),
    mean_income_prop = mean(income_prop, na.rm = TRUE),
    n = n()
  ) %>%
  arrange(year)

df.year <- left_join(yearlyTerritoryStats,
                     yearlyTerritoryFeatures,
                     by = c('year' = 'year'))

# scale covariates
df.year$mean_ruggedness_sc = scale(df.year$mean_ruggedness)[,1]      
df.year$average_gravel_km_sc = scale(df.year$average_gravel_km)[,1] 
df.year$average_paved_km_sc = scale(df.year$average_paved_km)[,1] 
df.year$pop_mean_sc = scale(df.year$pop_mean)[,1]            
df.year$mean_snow_sc = scale(df.year$mean_snow)[,1] 
df.year$mean_artificial_area_total_sc = scale(df.year$mean_artificial_area_total)[,1]
df.year$hunt_county2_sc = scale(df.year$hunt_county2)[,1]
df.year$mean_fi_sc = scale(df.year$mean_fi)[,1] 
df.year$mean_maxfi_sc = scale(df.year$mean_maxfi)[,1]
df.year$mean_number_neighbour_terr_sc = scale(df.year$mean_number_neighbour_terr)[,1] 
df.year$mean_bear_density_BZtiff_sc = scale(df.year$mean_bear_density_BZtiff)[,1]
#sum_wolfLKill_n = wolfLKill_n # count data - not scaled 
df.year$sum_wolfLKill_last5y_sc = scale(df.year$sum_wolfLKill_last5y)[,1]
#sum_sheepAttacks_n = sheepAttacks_n # count data - not scaled 
df.year$sum_NoAffectedSheep_last5y_sc = scale(df.year$sum_NoAffectedSheep_last5y)[,1] 
#sum_AttackDogs_n = AttackDogs_n # count data - not scaled 
df.year$sum_AttackDogs_last5y_sc = scale(df.year$sum_AttackDogs_last5y)[,1]
df.year$mean_IndividualIncome_sc = scale(df.year$mean_IndividualIncome)[,1] # count data - not scaled 
df.year$mean_income_prop_sc = scale(df.year$mean_income_prop)[,1] 
df.year$wolfPopSize_sc <- scale(df.year$wolfPopSize)[,1]

write_csv2(df.year,
          'modelInputData/df.year.csv')

```

### 4.2.1. Correlations

```{r}
# Note: covariate correlation had to be calculated for the different scales as 
# big discrepancies between scales were detected
colnames(df.year)
df.year2 <- df.year[2:24]

covToCorrelate_pop <- df.year2 %>%
  select(-c("disappearance", "n_territories_without_disap"))

cormat <- round(
  cor(
    covToCorrelate_pop,
    use = "pairwise.complete.obs",
    method = "spearman"),
  2)

cormat
upper_tri <- get_upper_tri(cormat)
melted_cormat <- reshape2::melt(upper_tri, na.rm = TRUE)

p1_pop <- ggplot(melted_cormat, aes(Var2, Var1, fill = value))+
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  ggtitle(label = "Spearman correlation - population level - all covariates") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1,
                                   hjust = 1, size = 13)) +
  theme(axis.text.y = element_text(size = 13)) +
  geom_text(aes(Var2, Var1, label = value), color = "white", size = 4.5)
p1_pop  

## correlation between metrics for the same variable ---------------------------

VarsToTest <- c("sum_AttackDogs_last5y",
                "sum_AttackDogs_n") # keep sum_AttackDogs_last5y

VarsToTest <- c("sum_sheepAttacks_n",
                "sum_NoAffectedSheep_last5y") # keep sum_NoAffectedSheep_last5

VarsToTest <- c("sum_wolfLKill_n",
                "sum_wolfLKill_last5y") # keep sum_wolfLKill_last5y

VarsToTest <- c("average_gravel_km",
                "average_paved_km") # keep both

VarsToTest <- c("mean_IndividualIncome",
                "mean_income_prop") # keep mean_income_prop

VarsToTest <- c("mean_fi",
                "mean_maxfi") # keep mean_fi

subset <- covToCorrelate %>%
  select(all_of(VarsToTest))

sub_cormat <- round(
  cor(
    subset,
    use = "pairwise.complete.obs",
    method = "spearman"),2
)

sub_upper_tri <- get_upper_tri(sub_cormat)
sub_melted_cormat <-
  reshape2::melt(sub_upper_tri, na.rm = TRUE)

p2_sub <- ggplot(sub_melted_cormat, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  ggtitle(label = "Spearman correlation - Population level") +
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1,
    size = 16
  )) +
  theme(axis.text.y = element_text(size = 16)) +
  geom_text(aes(Var2, Var1, label = value),
            color = "white",
            size = 6)
p2_sub

# subset the covariates to keep
subset_keep <- covToCorrelate_pop[c("wolfPopSize","mean_ruggedness",
                                "average_gravel_km", "average_paved_km",
                                "mean_artificial_area_total","pop_mean",
                                "mean_snow","sum_AttackDogs_last5y",
                                "sum_NoAffectedSheep_last5y",
                                "sum_wolfLKill_last5y", "hunt_county2",
                                "mean_income_prop","mean_bear_density_BZtiff",
                                "mean_number_neighbour_terr","mean_fi")]

# change covariates names
colnames(subset_keep) <- c("WolfPopSize","TerrRug","GravelRoad","PavedRoad",
                           "ArtificialArea","HumanDens","SnowCover",
                           "DogAttacks_last5y","SheepAttacks_last5y",
                           "LegWolfHarv_last5y","MooseHarv","IncomeProp",
                           "BearHarv","WolfNeighbourTerr","Mean_InbreedCoef")


## correlation between covariates representing the same mechanism --------------

VarsToTest2 <- c("TerrRug",
                 "GravelRoad",
                 "PavedRoad",
                 "ArtificialArea",
                 "HumanDens",
                 "SnowCover")

VarsToTest2 <- c("DogAttacks_last5y",
                 "SheepAttacks_last5y",
                 "LegWolfHarv_last5y",
                 "MooseHarv",
                 "IncomeProp")  

VarsToTest2 <- c("BearHarv",
                 "WolfNeighbourTerr")

sub_VarsToTest2 <- subset_keep %>%
  select(all_of(VarsToTest2))

sub_cormat2 <- round(
  cor(
    sub_VarsToTest2,
    use = "pairwise.complete.obs",
    method = "spearman"),2
)

sub_upper_tri2 <- get_upper_tri(sub_cormat2)
sub_melted_cormat2 <-
  reshape2::melt(sub_upper_tri2, na.rm = TRUE)

p3_sub <- ggplot(sub_melted_cormat2, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  ggtitle(label = "Spearman correlation - Population level - Socio-economic") +
  #change title depending on the hypothesis
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1,
    size = 16
  )) +
  theme(axis.text.y = element_text(size = 16)) +
  geom_text(aes(Var2, Var1, label = value),
            color = "white",
            size = 6)
p3_sub

## correlations between chosen variables ---------------------------------------

sub_cormat3 <- round(
  cor(
    subset_keep,
    use = "pairwise.complete.obs",
    method = "spearman"),2
)

sub_upper_tri3 <- get_upper_tri(sub_cormat3)
sub_melted_cormat3 <-
  reshape2::melt(sub_upper_tri3, na.rm = TRUE)

p4_sub <- ggplot(sub_melted_cormat3, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  ggtitle(label = "Spearman correlation - Population level - Chosen variables") +
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1,
    size = 16
  )) +
  theme(axis.text.y = element_text(size = 16)) +
  geom_text(aes(Var2, Var1, label = value),
            color = "white",
            size = 6)
p4_sub


## correlations of final covariates --------------------------------------------

subset_keep2 <- subset_keep[c("WolfPopSize","GravelRoad","SnowCover","TerrRug",
                              "HumanDens","Mean_InbreedCoef")]

sub_cormat4 <- round(
  cor(
    subset_keep2,
    use = "pairwise.complete.obs",
    method = "spearman"),2
)

sub_upper_tri4 <- get_upper_tri(sub_cormat4)
sub_melted_cormat4 <-
  reshape2::melt(sub_upper_tri4, na.rm = TRUE)

p5_sub <- ggplot(sub_melted_cormat4, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  ggtitle(label = "Spearman correlation - Population level - Final variables") +
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1,
    size = 16
  )) +
  theme(axis.text.y = element_text(size = 16)) +
  geom_text(aes(Var2, Var1, label = value),
            color = "white",
            size = 6)
p5_sub

```

### 4.2.2. Calculate min, max and mean

```{r}

min_cov <- format(sapply(df.year,min, na.rm = TRUE), scientific = FALSE)
min_cov <- as.data.frame(min_cov)

max_cov <- format(sapply(df.year,max, na.rm = TRUE), scientific = FALSE)
max_cov <- as.data.frame(max_cov)

mean_cov <- format(sapply(df.year,mean, na.rm = TRUE), scientific = FALSE)
mean_cov <- as.data.frame(mean_cov)

var <- data.frame(min_cov, max_cov, mean_cov)

write.csv(var, "info_variables_population.csv")

```

### 4.2.3. Plots of variables correlated with WolfPopSize

```{r}
# calculate coefficients for each variable

# format
max_cov_t <- as.data.frame(t(max_cov))
str(max_cov_t)
max_cov_t[c(1:42)] <- as.numeric(max_cov_t[c(1:42)])

# coefficients
coeff_attacksDogs <- max_cov_t$wolfPopSize/max_cov_t$sum_AttackDogs_last5y

coeff_attacksSheep <- max_cov_t$wolfPopSize/max_cov_t$sum_NoAffectedSheep_last5y

coeff_wolvesKilled <- max_cov_t$wolfPopSize/max_cov_t$sum_wolfLKill_last5y

coeff_moose <- max_cov_t$wolfPopSize/max_cov_t$hunt_county2

coeff_income <- max_cov_t$wolfPopSize/max_cov_t$mean_IndividualIncome

coeff_bear <- max_cov_t$wolfPopSize/max_cov_t$mean_bear_density_BZtiff

coeff_wolfNeigh <- max_cov_t$wolfPopSize/max_cov_t$mean_number_neighbour_terr
```


```{r}
# plots
# DISCRETE
plot_wolvesKilled <- 
  ggplot(df.year)  + 
  scale_linetype_manual("", values = 1) +
  geom_col(aes(x=year, y=sum_wolfLKill_last5y, group = 1, fill = "#660000"))+
  geom_line(aes(x=year, y=wolfPopSize / coeff_wolvesKilled, linetype = "Wolf population size"), colour = "black")+
  scale_y_continuous(name = "Number of wolves killed",
                     sec.axis = sec_axis(~.*coeff_wolvesKilled, name = "Wolf population size", 
                                         breaks = seq(0, 80, by = 10))) +
  scale_fill_manual('', values = '#660000', label = 'Number of wolves legally killed (5 years)') +
  theme_light() +
  theme(
    axis.title.y = element_text(size=14),
    axis.title.y.right = element_text(size=14),
    legend.position = "bottom",
    legend.box.margin=margin(-12,0,-5,0),
    legend.text=element_text(size=12)
  ) +
  xlab("Year")

###
plot_dogs <- 
  ggplot(df.year)  + 
  scale_linetype_manual("", values = 1) +
  geom_col(aes(x=year, y=sum_AttackDogs_last5y, group = 1, fill = "tan2"))+
  geom_line(aes(x=year, y=wolfPopSize / coeff_attacksDogs, linetype = "Wolf population size"), colour = "black")+
  scale_y_continuous(name = "Number of dogs attacked",
                     sec.axis = sec_axis(~.*coeff_attacksDogs, name = "Wolf population size",
                                         breaks = seq(0, 80, by = 10))) +
  scale_fill_manual('', values = 'tan2', label = 'Number of dogs attacked (5 years)') +
  theme_light() +
  theme(
    axis.title.y = element_text(size=14),
    axis.title.y.right = element_text(size=14),
    legend.position = "bottom",
    legend.box.margin=margin(-12,0,-5,0),
    legend.text=element_text(size=12)
  ) +
  xlab("Year")

###
plot_sheep <- 
  ggplot(df.year)  + 
  scale_linetype_manual("", values = 1) +
  geom_col(aes(x=year, y=sum_NoAffectedSheep_last5y, group = 1, fill = "sienna3"))+
  geom_line(aes(x=year, y=wolfPopSize / coeff_attacksSheep, linetype = "Wolf population size"), colour = "black")+
  scale_y_continuous(name = "Number of sheep attacked",
                     sec.axis = sec_axis(~.*coeff_attacksSheep, name = "Wolf population size",
                                         breaks = seq(0, 80, by = 10))) +
  scale_fill_manual('', values = 'sienna3', label = 'Number of sheep attacked (5 years)') +
  theme_light() +
  theme(
    axis.title.y = element_text(size=14),
    axis.title.y.right = element_text(size=14),
    legend.position = "bottom",
    legend.box.margin=margin(-12,0,-5,0),
    legend.text=element_text(size=12)
  ) +
  xlab("Year")

plot_grid(plot_wolvesKilled,plot_sheep,plot_dogs)

###
# CONTINUOUS

plot_income <- ggplot(df.year, aes(x = year)) +
  geom_line(aes(y = mean_IndividualIncome, colour = "Average annual income per municipality"), linewidth = 0.8) +
  geom_line(aes(y = wolfPopSize / coeff_income, colour = "Wolf population size"), linewidth = 0.8) +
  scale_color_manual(values = c("#006600", "black")) +
  scale_y_continuous(name = "Individual income",
                     sec.axis = sec_axis(~.*coeff_income, name = "Wolf population size",
                                         breaks = seq(0, 80, by = 10))) +
  theme_light() +
  theme(
    axis.title.y = element_text(size=14),
    axis.title.y.right = element_text(size=14),
    legend.position = "bottom",
    legend.box.margin=margin(-12,0,-5,0),
    legend.text=element_text(size=12)
  ) +
  labs(x = "Year",
       colour = "")


###
plot_bear <- ggplot(df.year, aes(x = year)) +
  geom_line(aes(y = mean_bear_density_BZtiff, colour = "Bear harvest (number/km2)"), linewidth = 0.8) +
  geom_line(aes(y = wolfPopSize / coeff_bear, colour = "Wolf population size"), linewidth = 0.8) +
  scale_color_manual(values = c("#0066CC", "black")) +
  scale_y_continuous(name = "Bear harvest (number/km2)",
                     sec.axis = sec_axis(~.*coeff_bear, name = "Wolf population size",
                                         breaks = seq(0, 80, by = 10))) +
  theme_light() +
  theme(
    axis.title.y = element_text(size=14),
    axis.title.y.right = element_text(size=14),
    legend.position = "bottom",
    legend.box.margin=margin(-12,0,-5,0),
    legend.text=element_text(size=12)
  ) +
  labs(x = "Year",
       colour = "")

###
plot_moose <- ggplot(df.year, aes(x = year)) +
  geom_line(aes(y = hunt_county2, colour = "Moose harvest (number/1000ha)"), linewidth = 0.8) +
  geom_line(aes(y = wolfPopSize / coeff_moose, colour = "Wolf population size"), linewidth = 0.8) +
  scale_color_manual(values = c("#FF9999", "black")) +
  scale_y_continuous(name = "Moose harvest",
                     sec.axis = sec_axis(~.*coeff_moose, name = "Wolf population size",
                                         breaks = seq(0, 80, by = 10))) +
  theme_light() +
  theme(
    axis.title.y = element_text(size=14),
    axis.title.y.right = element_text(size=14),
    legend.position = "bottom",
    legend.box.margin=margin(-12,0,-5,0),
    legend.text=element_text(size=12)
  ) +
  labs(x = "Year",
       colour = "")

###
plot_wolfNeigh <- ggplot(df.year, aes(x = year)) +
  geom_line(aes(y = mean_number_neighbour_terr, colour = "Neighboring wolf territories"), linewidth = 0.8) +
  geom_line(aes(y = wolfPopSize / coeff_moose, colour = "Wolf population size"), linewidth = 0.8) +
  scale_color_manual(values = c("#9966CC", "black")) +
  scale_y_continuous(name = "Number of neighboring wolf territories",
                     sec.axis = sec_axis(~.*coeff_wolfNeigh, name = "Wolf population size",
                                         breaks = seq(0, 80, by = 10))) +
  theme_light() +
  theme(
    axis.title.y = element_text(size=14),
    axis.title.y.right = element_text(size=14),
    legend.position = "bottom",
    legend.box.margin=margin(-12,0,-5,0),
    legend.text=element_text(size=12)
  ) +
  labs(x = "Year",
       colour = "")

###
plot_grid(plot_income, plot_moose, plot_bear, plot_wolfNeigh, nrow=2)

###
plot_grid(plot_wolvesKilled,plot_sheep,plot_dogs,plot_income, plot_moose, plot_bear, plot_wolfNeigh, nrow=4)



```


## 4.3. Municipality level

```{r}
## Model input data - Municipality level
# summarise territory data per municipality ------------------------------------
municipalityTerritoryStats <-
  dataToModel8 %>%
  group_by(territory, Spring) %>%
  slice_max(mean_fi, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(territory, Spring) %>%
  dplyr::mutate(year = Spring, .before = mod_id) %>%
  group_by(NAME_2) %>%
  dplyr::summarise(
    wolfPopSize = length(unique(territory)),
    disappearance = sum(Fate == "illegal", na.rm = TRUE),
    n_territories_without_disap = wolfPopSize - disappearance,
    prop_disappearance = disappearance / wolfPopSize,
  ) 

#head(municipalityTerritoryStats, n=30)
# proportion of disappearance per municipality
p <-
  ggplot(municipalityTerritoryStats,
         aes(x = reorder(NAME_2,+prop_disappearance),
             y = prop_disappearance)) +
  geom_bar(stat = "identity") +
  theme(axis.text.x = element_text(angle = 90))
p

municipalityTerritoryFeatures <-
  territoriesToModel %>%
  dplyr::mutate(year = Spring, .before = mod_id) %>%
  #dplyr::filter(Fate_binary == 1) %>% # filter only disappearances
  dplyr::group_by(NAME_2) %>%
  dplyr::summarise(
    mean_ruggedness = mean(ruggedness_mean, na.rm = TRUE),
    average_gravel_km = mean(average_gravel_km, na.rm = TRUE),
    average_paved_km = mean(average_paved_km, na.rm = TRUE),
    pop_mean = mean(pop_mean, na.rm = TRUE),
    mean_snow = mean(mean_snow, na.rm = TRUE),
    mean_artificial_area_total = mean(artificial_area_total, na.rm = TRUE),
    hunt_county2 = mean(hunt_county2, na.rm = TRUE),
    mean_fi = mean(mean_fi, na.rm = TRUE),
    mean_maxfi = mean(max_fi, na.rm = TRUE),
    mean_number_neighbour_terr = mean(number_neighbour_terr, na.rm = TRUE),
    mean_bear_density_BZtiff = mean(bear_density_BZtiff, na.rm = TRUE),
    sum_wolfLKill_n = sum(wolfLKill_n, na.rm = TRUE),
    sum_wolfLKill_last5y = sum(wolfLKill_last5y, na.rm = TRUE),
    sum_sheepAttacks_n = sum(sheepAttacks_n, na.rm = TRUE),
    sum_NoAffectedSheep_last5y = sum(NoAffectedSheep_last5y, na.rm = TRUE),
    sum_AttackDogs_n = sum(AttackDogs_n, na.rm = TRUE),
    sum_AttackDogs_last5y = sum(AttackDogs_last5y, na.rm = TRUE),
    mean_IndividualIncome = mean(IndividualIncome, na.rm = TRUE),
    mean_income_prop = mean(income_prop, na.rm = TRUE),
    n = n()
  ) %>%
  arrange(NAME_2)

municipalityTerritoryStats <-
  municipalityTerritoryStats %>%
  dplyr::filter(
    NAME_2 %in% unique(municipalityTerritoryFeatures$NAME_2)
  )

df.municipality <-
  left_join(
    municipalityTerritoryStats,
    municipalityTerritoryFeatures,
    by = c('NAME_2' = 'NAME_2')
  )

view(df.municipality)
# double checking 
# head(df.municipality)
# table(df.municipality$sum_fate)
# sum((df.municipality$sum_fate))
# sum((df.municipality$n_territories))

# scale covariates
df.municipality$mean_ruggedness_sc = scale(df.municipality$mean_ruggedness)[,1]      
df.municipality$average_gravel_km_sc = scale(df.municipality$average_gravel_km)[,1] 
df.municipality$average_paved_km_sc = scale(df.municipality$average_paved_km)[,1] 
df.municipality$pop_mean_sc = scale(df.municipality$pop_mean)[,1]            
df.municipality$mean_snow_sc = scale(df.municipality$mean_snow)[,1] 
df.municipality$mean_artificial_area_total_sc = scale(df.municipality$mean_artificial_area_total)[,1]
df.municipality$hunt_county2_sc = scale(df.municipality$hunt_county2)[,1]
df.municipality$mean_fi_sc = scale(df.municipality$mean_fi)[,1] 
df.municipality$mean_maxfi_sc = scale(df.municipality$mean_maxfi)[,1]
df.municipality$mean_number_neighbour_terr_sc = scale(df.municipality$mean_number_neighbour_terr)[,1]
df.municipality$mean_bear_density_BZtiff_sc = scale(df.municipality$mean_bear_density_BZtiff)[,1]
#sum_wolfLKill_n = wolfLKill_n # count data - not scaled 
df.municipality$sum_wolfLKill_last5y_sc = scale(df.municipality$sum_wolfLKill_last5y)[,1]
#sum_sheepAttacks_n = sheepAttacks_n # count data - not scaled 
df.municipality$sum_NoAffectedSheep_last5y_sc = scale(df.municipality$sum_NoAffectedSheep_last5y)[,1]
#sum_AttackDogs_n = AttackDogs_n # count data - not scaled 
df.municipality$sum_AttackDogs_last5y_sc = scale(df.municipality$sum_AttackDogs_last5y)[,1] 
df.municipality$mean_IndividualIncome_sc = scale(df.municipality$mean_IndividualIncome)[,1] # count data - not scaled 
df.municipality$mean_income_prop_sc = scale(df.municipality$mean_income_prop)[,1]
df.municipality$wolfPopSize_sc = scale(df.municipality$wolfPopSize)[,1]

write.csv2(df.municipality,
           'modelInputData/df.municipality.csv')
```

### 4.3.1. Correlations 

```{r}
# Note: covariate correlation had to be calculated for the different scales as big discrepancies between scales were detected
colnames(df.municipality)
df.municipality2 <- df.municipality[2:24]

covToCorrelate_mun <- df.municipality2 %>%
  select(-c("disappearance", "n_territories_without_disap"))

cormat <- round(
  cor(
    covToCorrelate_mun,
    use = "pairwise.complete.obs",
    method = "spearman"),
  2)

cormat
upper_tri <- get_upper_tri(cormat)
melted_cormat <- reshape2::melt(upper_tri, na.rm = TRUE)

p1_mun <- ggplot(melted_cormat, aes(Var2, Var1, fill = value))+
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  ggtitle(label = "Spearman correlation - continuous variables") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1,
                                   hjust = 1, size = 13)) +
  theme(axis.text.y = element_text(size = 13)) +
  geom_text(aes(Var2, Var1, label = value), color = "white", size = 4.5)
p1_mun  

# highlyCorrelatedVars <- melted_cormat %>%
#   filter(value >0.5 | value < -0.5)
# #highlyCorrelatedVars

## correlation between metrics for the same variable ---------------------------
VarsToTest <- c("sum_AttackDogs_last5y",
                "sum_AttackDogs_n") # keep sum_AttackDogs_last5y

VarsToTest <- c("sum_sheepAttacks_n",
                "sum_NoAffectedSheep_last5y") # keep sum_NoAffectedSheep_last5y

VarsToTest <- c("sum_wolfLKill_n",
                "sum_wolfLKill_last5y") # keep sum_wolfLKill_last5y

VarsToTest <- c("average_gravel_km",
                "average_paved_km") # keep average_gravel_km

VarsToTest <- c("mean_IndividualIncome",
                "mean_income_prop") # keep mean_income_prop

VarsToTest <- c("mean_fi",
                "mean_maxfi") # keep mean_fi

subset <- covToCorrelate %>%
  select(all_of(VarsToTest))

sub_cormat <- round(
  cor(
    subset,
    use = "pairwise.complete.obs",
    method = "spearman"),2
)

sub_upper_tri <- get_upper_tri(sub_cormat)
sub_melted_cormat <-
  reshape2::melt(sub_upper_tri, na.rm = TRUE)

p2_sub <- ggplot(sub_melted_cormat, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  ggtitle(label = "Spearman correlation - Population level") +
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1,
    size = 16
  )) +
  theme(axis.text.y = element_text(size = 16)) +
  geom_text(aes(Var2, Var1, label = value),
            color = "white",
            size = 6)
p2_sub

# subset the covariates to keep
subset_keep <- covToCorrelate_mun[c("wolfPopSize","mean_ruggedness",
                                "average_gravel_km",
                                "mean_artificial_area_total",
                                "pop_mean","mean_snow","sum_AttackDogs_last5y",
                                "sum_NoAffectedSheep_last5y",
                                "sum_wolfLKill_last5y",
                                "hunt_county2","mean_income_prop",
                                "mean_bear_density_BZtiff",
                                "mean_number_neighbour_terr",
                                "mean_fi")]

# change covariates names
colnames(subset_keep) <- c("WolfPopSize","TerrRug","GravelRoad",
                           "ArtificialArea","HumanDens","SnowCover",
                           "DogAttacks_last5y","SheepAttacks_last5y",
                           "LegWolfHarv_last5y","MooseHarv","IncomeProp",
                           "BearHarv","WolfNeighbourTerr","Mean_InbreedCoef")

## correlation between covariates representing the same mechanism --------------
VarsToTest2 <- c("TerrRug",
                 "GravelRoad",
                 "ArtificialArea",
                 "HumanDens",
                 "SnowCover")

VarsToTest2 <- c("DogAttacks_last5y",
                 "SheepAttacks_last5y",
                 "LegWolfHarv_last5y",
                 "MooseHarv",
                 "IncomeProp")  

VarsToTest2 <- c("BearHarv",
                 "WolfNeighbourTerr")

sub_VarsToTest2 <- subset_keep %>%
  select(all_of(VarsToTest2))

sub_cormat2 <- round(
  cor(
    sub_VarsToTest2,
    use = "pairwise.complete.obs",
    method = "spearman"),2
)

sub_upper_tri2 <- get_upper_tri(sub_cormat2)
sub_melted_cormat2 <-
  reshape2::melt(sub_upper_tri2, na.rm = TRUE)

p3_sub <- ggplot(sub_melted_cormat2, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  ggtitle(label = "Spearman correlation - Municipality level - Socio-economic") +
  #change title depending on the hypothesis
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1,
    size = 16
  )) +
  theme(axis.text.y = element_text(size = 16)) +
  geom_text(aes(Var2, Var1, label = value),
            color = "white",
            size = 6)
p3_sub

## correlations between chosen variables ---------------------------------------

sub_cormat3 <- round(
  cor(
    subset_keep,
    use = "pairwise.complete.obs",
    method = "spearman"),2
)

sub_upper_tri3 <- get_upper_tri(sub_cormat3)
sub_melted_cormat3 <-
  reshape2::melt(sub_upper_tri3, na.rm = TRUE)

p4_sub <- ggplot(sub_melted_cormat3, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  ggtitle(label = "Spearman correlation - Municipality level - Chosen variables") +
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1,
    size = 16
  )) +
  theme(axis.text.y = element_text(size = 16)) +
  geom_text(aes(Var2, Var1, label = value),
            color = "white",
            size = 6)
p4_sub


## correlations of final covariates --------------------------------------------

subset_keep2 <- subset_keep[c("WolfPopSize","TerrRug","MooseHarv",
                              "IncomeProp","WolfNeighbourTerr","Mean_InbreedCoef")]

sub_cormat4 <- round(
  cor(
    subset_keep2,
    use = "pairwise.complete.obs",
    method = "spearman"),2
)

sub_upper_tri4 <- get_upper_tri(sub_cormat4)
sub_melted_cormat4 <-
  reshape2::melt(sub_upper_tri4, na.rm = TRUE)

p5_sub <- ggplot(sub_melted_cormat4, aes(Var2, Var1, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c() +
  ggtitle(label = "Spearman correlation - Municipality level - Final variables") +
  theme_minimal() +
  theme(axis.text.x = element_text(
    angle = 45,
    vjust = 1,
    hjust = 1,
    size = 16
  )) +
  theme(axis.text.y = element_text(size = 16)) +
  geom_text(aes(Var2, Var1, label = value),
            color = "white",
            size = 6)
p5_sub
```

### 4.3.2. Calculate min, max and mean

```{r}

min_cov <- format(sapply(df.municipality,min, na.rm = TRUE), scientific = FALSE)
min_cov <- as.data.frame(min_cov)

max_cov <- format(sapply(df.municipality,max, na.rm = TRUE), scientific = FALSE)
max_cov <- as.data.frame(max_cov)

mean_cov <- format(sapply(df.municipality,mean, na.rm = TRUE), scientific = FALSE)
mean_cov <- as.data.frame(mean_cov)

var <- data.frame(min_cov, max_cov, mean_cov)

write.csv(var, "info_variables_municipality.csv")

```


